### 2018-09-08 ESUG2018/esug.tex
Add src for ESUG2018 presentation.

\documentclass{beamer}

\usepackage{euler}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{textcomp}
\usepackage{minted}
\usepackage[scaled=0.8]{beramono}

\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}

\usefonttheme[onlymath]{serif}

\usemintedstyle{friendly}

\setbeamertemplate{blocks}[rounded]

%Information to be included in the title page:
\title{Relational Programming in Smalltalk}
\author{Massimo Nocentini}
\institute{University of Florence, Italy}
\date{ESUG2018}

\begin{document}

\frame{\titlepage}

\begin{frame}[fragile]
\frametitle{}
\begin{minted}[fontsize=\small]{smalltalk}
outline
    ^ LinkedList new
        add: 'Resolution by refutation';
        add: 'Resolution by unification';
        yourself
\end{minted}
\end{frame}

\begin{frame}[fragile]

\frametitle{Resolution by \textit{Refutation}}

Let $\alpha$ be a sentence in \textit{CNF} and $M(\alpha)$ the set of models that
satisfy it, where a model is a set of assignments that make $\alpha$ true.
\vfill
$\alpha$ is \textit{valid} if it is true in \textit{all} models; oth,\\
$\alpha$ is \textit{satisfiable} if it is true in \textit{some} model.
\vfill
\textit{Logical reasoning} boils down to decide the \textit{entailment}
relation $\models$
\begin{displaymath}
\begin{split}
M(\alpha) &\subseteq M(\beta) \leftrightarrow \\
\alpha &\models \beta \leftrightarrow \\
(\alpha &\Rightarrow \beta) \text{ is valid } \leftrightarrow\\
\neg(\neg\alpha &\vee \beta) \text{ is unsatisfiable}
\end{split}
\end{displaymath}
therefore, to prove a sentence $\alpha$ reduces to decide
$\neg\alpha\models\perp$,\\where $\perp$ is the (\textit{unsatisfiable}) empty
clause.
\end{frame}

\begin{frame}[fragile]
\frametitle{Resolution by \textit{Refutation}}
The \textit{resolution rule} is a \textit{complete} inference algorithm; \\
let $l_{q}, m_{w} \in\lbrace 0,1\rbrace$ in
\begin{displaymath}
{\left(l_{0}, \ldots, l_{i}, \ldots, l_{j-1}\right) \quad \left(m_{0}, \ldots, m_{r},\ldots, m_{k-1}\right) \quad l_{i} = \neg m_{r}
\over
\left(l_{0},\ldots, l_{i-1},l_{i+1}, \ldots,l_{j-1}, m_{0},\ldots, m_{r-1},m_{r+1},\ldots, m_{k-1}\right)}
\end{displaymath}
where $\left(l_{0},\ldots, l_{i}, \ldots, l_{j-1}\right) = l_{0}\vee \ldots \vee l_{i} \vee \ldots \vee l_{j-1}$.
\vfill
The \textit{DPLL} algorithm is a recursive, depth-first enumeration of models
using the resolution rule, paired with heuristics \textit{early termination},
\textit{pure symbol} and \textit{unit clause} to speed up.
\end{frame}

\begin{frame}[fragile]
\frametitle{Resolution by \textit{Unification}}

\textit{Unification} is the process of solving \textit{equations among symbolic
expressions}; a \textit{solution} is denoted as a \textit{substitution}
$\theta$, namely a mapping that assigns a symbolic values to free variables.
\vfill
Let $x$ and $y$ be free variables, the set
$$\lbrace cons(x,cons(x,nil)) = cons(2,y)\rbrace$$
has solution $\theta = \lbrace x \mapsto 2, y \mapsto cons(2,nil) \rbrace$;
moreover, the set
$$ \lbrace y = cons(2,y) \rbrace $$
has no \textit{finite} solution; oth, $\theta = \lbrace y \mapsto
cons(2,cons(2,cons(2,...))) \rbrace$ is a solution by \textit{bisimulation}.
\end{frame}

\begin{frame}[fragile]
\frametitle{Resolution by \textit{Unification}}
let $G$ be a set of equations, unification rules are
\begin{description}
\item[delete] $G \cup \lbrace t = t \rbrace \rightarrow G$
\item[decompose] $G \cup \lbrace f(s_{0}, \ldots, s_{k}) = f(t_{0}, \ldots, t_{k})\rbrace$ entails
$$G \cup \lbrace s_{0}=t_{0},\ldots, s_{k}=t_{k} \rbrace$$
\item[conflict] if $f\neq g \vee k\neq m$ then $$G \cup \lbrace f(s_{0}, \ldots, s_{k}) = g(t_{0}, \ldots, t_{m})\rbrace \rightarrow \,\perp$$
\item[eliminate] if $x \not\in vars(t)$ and $x \in vars(G)$ then $$G \cup \lbrace x = t\rbrace \rightarrow G\lbrace x \mapsto t\rbrace \cup \left\lbrace x \triangleq t\right\rbrace $$
\item[occur check] if $x \in vars(f(s_{0},\ldots,s_{k}))$ then $$G \cup \lbrace x = t(s_{0}, \ldots, s_{k}\rbrace \rightarrow \,\perp$$
\end{description}
%without *occur checks*, generating a substitution $\theta$ is a *recursive enumerable* problem
\end{frame}

\begin{frame}[fragile]
\frametitle{microkanren}
\begin{itemize}
\item a \textit{DSL} for relational programming written in Scheme
\item \textit{purely functional} implementation of \textit{miniKanren}
\item \textit{explicit streams} of satisfying states, \textit{goal-based} approach
\item resolution by unification via \textit{structural induction}
\item \textit{complete}, but \textit{unfair}, search strategy
\end{itemize}

\iffalse
my contribution
- _Pythonic_ [implementation][mkpy]: functional at the core, objective at the interface
- generators subsume _countably_-satisfiable relations; complete, _fair_ [search][dovetail]
- _The Reasoned Schemer_ fully tested via [Travis CI][travis]; moreover, [read the docs][rtfd]
- case studies: Smullyan puzzles and combinatorics
- tweaking HOL Light for _certified deductions_, [wip][klight]
\fi
\end{frame}

\begin{frame}[fragile]
\frametitle{Chain subclass: \#Bottom}

\begin{minted}[fontsize=\footnotesize]{smalltalk}
BlockClosure>>links: anObj
    ^ Chain item: anObj linker: self
\end{minted}
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Chain class>>bottom
    ^ Bottom new
\end{minted}
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Chain class>>item: anObj linker: aBlockClosure
    ^ Knot new 
        item: anObj;
        linker: aBlockClosure;
        yourself
\end{minted}
\begin{minted}[fontsize=\footnotesize]{smalltalk}
bind: aGoal interleaved: anInterleaved
    ^ self
\end{minted}
\begin{minted}[fontsize=\footnotesize]{smalltalk}
mplus: anotherChain interleaved: anInterleaved
    ^ anotherChain value
\end{minted}
\begin{minted}[fontsize=\footnotesize]{smalltalk}
atMost: anInteger
    ^ self
\end{minted}
\begin{minted}[fontsize=\footnotesize]{smalltalk}
mature
    ^ LinkedList new
\end{minted}
\end{frame}

\begin{frame}[fragile]
\frametitle{Chain subclass: \#Knot}
\begin{minted}[fontsize=\footnotesize]{smalltalk}
bind: aGoal interleaved: anInterleaved
    | alpha beta |
    alpha := aGoal onState: item.
    beta := [ self next bind: aGoal interleaved: anInterleaved ].
    ^ alpha mplus: beta interleaved: anInterleaved
\end{minted}
\begin{minted}[fontsize=\footnotesize]{smalltalk}
mplus: anotherChain interleaved: anInterleaved
    ^ [ :_ | anotherChain value 
                mplus: [ self next ] 
                interleaved: anInterleaved ] links: item
\end{minted}
\begin{minted}[fontsize=\footnotesize]{smalltalk}
next
    ^ linker value: item
\end{minted}
\begin{minted}[fontsize=\footnotesize]{smalltalk}
atMost: n
    ^ n isZero
        ifTrue: [ Chain bottom ]
        ifFalse: [ [ :_ | self next value atMost: n - 1 ] links: item ]
\end{minted}
\begin{minted}[fontsize=\footnotesize]{smalltalk}
mature
    ^ self next mature
        addFirst: item;
        yourself
\end{minted}
\end{frame}

\begin{frame}[fragile]
\frametitle{ChainTest}
\begin{minted}[fontsize=\footnotesize]{smalltalk}
ints: i
    ^ [ :a | self ints: a + 1 ] links: i
\end{minted}
\begin{minted}[fontsize=\footnotesize]{smalltalk}
fib: m fib: n
    ^ [ :_ | self fib: n fib: m + n ] links: m
\end{minted}
\begin{minted}[fontsize=\footnotesize]{smalltalk}
collatz: o
    ^ [ :_ | o even
                ifTrue: [ self collatz: o / 2 ]
                ifFalse: [ self collatz: 3 * o + 1 ] ] links: o
\end{minted}
\begin{minted}[fontsize=\footnotesize]{smalltalk}
testNumbers
    self assert: (self nats atMost: 10) mature equals: (0 to: 9).
    self
        assert: (self fibs atMost: 10) mature
        equals: {0 . 1 . 1 . 2 . 3 . 5 . 8 . 13 . 21 . 34}.
    self
        assert: ((self collatz: 10) atMost: 10) mature
        equals: {10 . 5 . 16 . 8 . 4 . 2 . 1 . 4 . 2 . 1}.
    self
        assert: ((self gcd: 18 gcd: 32) atMost: 10) mature
        equals: {18 . 32 . 18 . 14 . 4 . 2 . 2}
\end{minted}
\end{frame}

\begin{frame}[fragile]
\frametitle{Chain combinators}
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Bind class>>combine: aGoal with: aCollection
    ^ self new
        combiner: aGoal;
        stream: aCollection;
        yourself
\end{minted}
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Bind>>interleavedStrategy: anInterleaved
    ^ stream bind: combiner interleaved: anInterleaved
\end{minted}
\begin{minted}[fontsize=\footnotesize]{smalltalk}
MPlus class>>with: aCollection with: anotherCollection
    ^ self new
        left: aCollection;
        right: anotherCollection;
        yourself
\end{minted}
\begin{minted}[fontsize=\footnotesize]{smalltalk}
MPlus>>interleavedStrategy: anInterleaved
    ^ left mplus: right interleaved: anInterleaved
\end{minted}
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Sequential>>of: aStreamCombination
    ^ aStreamCombination sequentialStrategy: self
\end{minted}
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Interleaved>>of: aStreamCombination
    ^ aStreamCombination interleavedStrategy: self
\end{minted}
\end{frame}

\begin{frame}[fragile]
\frametitle{Goals}
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Succeed>>onState: aState
    ^ Chain with: aState
\end{minted}
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Fail>>onState: aState
    ^ Chain bottom
\end{minted}
\begin{minted}[fontsize=\footnotesize]{smalltalk}
State>>at: aVar put: aValue
    | s |
    s := substitution copy.
    s
        at: aVar
        ifPresent: [ :v | 
            aValue = v
                ifFalse: [ UnificationError signal ] ]
        ifAbsent: [ s at: aVar put: aValue ].
    ^ self class new
        birthdate: birthdate;
        substitution: s;
        yourself
\end{minted}
\end{frame}

\begin{frame}[fragile]
\frametitle{Goals}
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Fresh>>onState: aState
    ^ aState collectVars: (1 to: receiver numArgs) forFresh: self
\end{minted}
\begin{minted}[fontsize=\footnotesize]{smalltalk}
State>>collectVars: aCollection forFresh: aFresh
    | nextState vars |
    nextState := self class new
        substitution: substitution;
        birthdate: birthdate + aCollection size;
        yourself.
    vars := aCollection collect: [ :i | Var id: i ].
    ^ aFresh onState: nextState withVars: vars
\end{minted}
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Fresh>>onState: aState withVars: aCollection
    | g |
    vars := aCollection.
    g := receiver valueWithArguments: vars.
    ^ g onState: aState
\end{minted}
\begin{minted}[fontsize=\footnotesize]{smalltalk}
BlockClosure>>fresh
    ^ Goal fresh: self
\end{minted}
\end{frame}

\begin{frame}[fragile]
\frametitle{Goals}
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Object>>unifyWith: another
    ^ Goal unify: self with: another
\end{minted}
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Unify>>onState: aState
    ^ [ | extended_state |
        extended_state := Unifier new 
                            unify: this with: that onState: aState.
        Goal succeed onState: extended_state ]
            on: UnificationError
            do: [ Goal fail onState: aState ]
\end{minted}
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Unifier>>unify: anObj with: anotherObj onState: aState
    | aWalkedObj anotherWalkedObj |
    aWalkedObj := aState walk: anObj.
    anotherWalkedObj := aState walk: anotherObj.
    ^ aWalkedObj unifyWith: anotherWalkedObj 
                 usingUnifier: self 
                 onState: aState
\end{minted}
\begin{minted}[fontsize=\footnotesize]{smalltalk}
State>>walk: anObj
    | k |
    k := anObj.
    [ k := substitution at: k ifAbsent: [ ^ k ] ] repeat
\end{minted}
\end{frame}

\begin{frame}[fragile]
\frametitle{Unifier}
\begin{minted}[fontsize=\footnotesize]{smalltalk}
unifyObject: anObj withObject: anotherObj onState: aState
    ^ anObj = anotherObj
        ifTrue: [ aState ]
        ifFalse: [ UnificationError signal ]
\end{minted}
\begin{minted}[fontsize=\footnotesize]{smalltalk}
unifyVar: aVar withObject: anObject onState: aState
    ^ aState at: aVar put: anObject
\end{minted}
\begin{minted}[fontsize=\footnotesize]{smalltalk}
unifyVar: aVar withVar: anotherVar onState: aState
    ^ aVar = anotherVar
        ifTrue: [ aState ]
        ifFalse: [ 
            self unifyVar: aVar withObject: anotherVar onState: aState ]
\end{minted}
\begin{minted}[fontsize=\footnotesize]{smalltalk}
unifyLinkedList: aCollection withLinkedList: anotherCollection onState: aState
    ^ aCollection size = anotherCollection size
        ifTrue: [ (aCollection zip: anotherCollection)
            inject: aState
            into: [ :s :pair | self unify: pair key 
                                    with: pair value 
                                    onState: s ] ]
        ifFalse: [ UnificationError signal ]
\end{minted}
\end{frame}

\begin{frame}[fragile]
\frametitle{Unifier}
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Disj>>onState: aState
    ^ interleaving of: ((either onState: aState) 
        mplus: [ or onState: aState ])
\end{minted}
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Conj>>onState: aState
    ^ interleaving of: ((both onState: aState) bind: and)
\end{minted}
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Cond>>if: ifGoal then: thenGoal
    clauses add: ifGoal -> thenGoal
\end{minted}
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Cond>>ifPure: aStrategy
    if := [ :clause :otherwise | 
            IfPure new
                question: clause key 
                    answer: clause value 
                    otherwise: otherwise;
                streamCombinationStrategy: aStrategy;
                yourself ]
\end{minted}
\end{frame}

\begin{frame}[fragile]
\frametitle{Unifier}
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Cond>>e
    self ifPure: Sequential new
\end{minted}
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Cond>>i
    self ifPure: Interleaved new
\end{minted}
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Cond>>onState: aState
    | g |
    else ifNil: [ self else: false asGoal ].
    g := clauses copy
            add: else;
            reduceRight: if.
    ^ g onState: aState
\end{minted}
\end{frame}

\begin{frame}[fragile]
\frametitle{Unifier}
Let $\mathcal{D}$ be the set of \textit{Dyck paths} and let $\leadsto$ be the \textit{CFG}
$$
\left\lbrace\begin{array}{l}
\leadsto = \varepsilon \\
\leadsto = ( \leadsto ) \leadsto \\
\end{array}\right\rbrace
$$
question: enumerate $\mathcal{D}$ using $\leadsto$.
\begin{minted}[fontsize=\footnotesize]{smalltalk}
dycko: alpha
    ^ Goal cond e
        if: alpha nilo then: true asGoal;
        else: [ :beta :gamma | 
            (sexpTheory let: alpha 
                        be: ($( cons: beta) 
                        append: ($) cons: gamma)) & 
            ([ self dycko: beta ] eta & 
             [ self dycko: gamma ] eta) ] fresh
\end{minted}
\end{frame}

\begin{frame}[fragile]
\frametitle{Unifier}
\begin{minted}[fontsize=\footnotesize]{smalltalk}
testDycko
    | g |

    "enumeration"
    g := [ :alpha | combTheory dycko: alpha ] fresh.
    self
        assert: (g solutions atMost: 20)
        equals:
            ({nil . '()' . '(())' . '()()' . '(()())' . '()(())' . 
            '(())()' . '()()()' . '(()()())'.  '()(()())' . 
            '(())(())' . '()()(())' . '((()))' . '()(())()' . 
            '(())()()' . '()()()()'.  '(()()()())' . '()(()()())' . 
            '(())(()())' . '()()(()())'} collect: #asCons).

    "an invalid Dyck path"
    g := [ :alpha | combTheory dycko: '(()(())()(' asCons ] fresh.  
    self assert: g solutions all equals: {}
\end{minted}
\end{frame}

\begin{frame}[fragile]
\frametitle{Unifier}
Let $X$ and $Y$ be natural numbers in machine 
\begin{displaymath}
\mathcal{C} =  \left \lbrace{ \over 2X \stackrel{\circ}{\rightarrow} X} , {X \stackrel{\circ}{\rightarrow} Y \over 3X \stackrel{\circ}{\rightarrow} Y 2 Y} \right \rbrace
\end{displaymath}
question: does exist a number $\alpha$ such that $ \alpha \stackrel{\circ}{\rightarrow} \alpha $?
\begin{minted}[fontsize=\footnotesize]{smalltalk}
consumes: two_alpha produces: alpha machine: aMachine
    ^ two_alpha unifyWith: (2 cons: alpha)
\end{minted}
\begin{minted}[fontsize=\footnotesize]{smalltalk}
consumes: three_alpha produces: alpha_two_alpha machine: aMachine
    ^ [ :beta :gamma | 
        (three_alpha unifyWith: (3 cons: beta)) &
        ((self associate: gamma is: alpha_two_alpha machine: aMachine) & 
        (aMachine proves: beta relates: gamma)) ] fresh
\end{minted}
\end{frame}

\begin{frame}[fragile]
\frametitle{Unifier}
\begin{minted}[fontsize=\footnotesize]{smalltalk}
InductiveRelationsTheory>>proves: anObj relates: anotherObj
    | g |
    g := Goal cond i.
    rules do: [ :r | g if: (r consumes: anObj 
                              produces: anotherObj 
                              machine: self) 
                       then: true asGoal ].
    ^ g
\end{minted}
\begin{minted}[fontsize=\footnotesize]{smalltalk}
testFirstMachine
    | g |

    "McCulloch's first machine"
    g := [ :a | self mcculloch proves: a relates: a ] fresh.
    self assert: (g solutions atMost: 1) equals: {#(3 2 3) asCons}.

    "Montecarlo lock"
    g := [ :a | self mclock proves: a relates: a ] fresh.
    self
        assert: ((g solutions atMost: 1) collect: #asLinkedList)
        equals: {#(5 4 6 4 2 5 4 6 4 2)}
\end{minted}
\end{frame}

\end{document}

### 2018-09-08 ESUG2018/esug.tex
Add some slides.

@@
 \end{minted}
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Main idea}
+\begin{displaymath}
+\forall a,b,c.\,P(a,b,c) \leftrightarrow a + b = c \quad\text{entails}\quad P(1,2,3)
+\end{displaymath}
+can be expressed using either the \textit{imperative style}
+\begin{minted}[fontsize=\footnotesize]{smalltalk}
+a := 1.
+b := 2.
+c := a + b.
+Object assert: [ c = 3 ].
+\end{minted}
+or the \textit{functional style}
+\begin{minted}[fontsize=\footnotesize]{smalltalk}
+([ :a :b | a + b ] value: 1 value: 2) = 3
+\end{minted}
+or, finally, the \textit{declarative style}
+\begin{minted}[fontsize=\footnotesize]{smalltalk}
+[ :a :b :c | a + b = c ] value: 1 value: 2 value: 3
+\end{minted}
+
 \end{frame}
@@
 \item[eliminate] if $x \not\in vars(t)$ and $x \in vars(G)$ then $$G \cup \lbrace x = t\rbrace \rightarrow G\lbrace x \mapsto t\rbrace \cup \left\lbrace x \triangleq t\right\rbrace $$
-\item[occur check] if $x \in vars(f(s_{0},\ldots,s_{k}))$ then $$G \cup \lbrace x = t(s_{0}, \ldots, s_{k}\rbrace \rightarrow \,\perp$$
+\item[occur check] if $x \in vars(f(s_{0},\ldots,s_{k}))$ then $$G \cup \lbrace x = t(s_{0}, \ldots, s_{k})\rbrace \rightarrow \,\perp$$
 \end{description}
@@
 \end{itemize}
+
+A \textit{goal} is an object that responds to the \verb|#onState:| selector, it
+receives a \textit{substitution} and returns a \verb|Chain| object of
+substitutions.
 
@@
 $$
-\left\lbrace\begin{array}{l}
-\leadsto = \varepsilon \\
-\leadsto = ( \leadsto ) \leadsto \\
-\end{array}\right\rbrace
+\leadsto = \varepsilon \,\left|\, ( \leadsto ) \leadsto \right.
 $$
-question: enumerate $\mathcal{D}$ using $\leadsto$.
+where $\varepsilon$ is the empty string; question: enumerate $\mathcal{D}$ using $\leadsto$.
 \begin{minted}[fontsize=\footnotesize]{smalltalk}

### 2018-09-13 ESUG2018/esug.tex
ESUG2018 presentation.

--- 2018-09-08 ESUG2018/esug.tex
@@
 \usepackage[scaled=0.8]{beramono}
-
+\usepackage{CJKutf8}
 \usepackage{amsmath}
@@
     ^ LinkedList new
-        add: 'Resolution by refutation';
-        add: 'Resolution by unification';
+        add: 'me and motivations';
+        add: 'Refutation and Unification resolutions ';
+        add: 'Microkanren in Smalltalk';
+        add: 'Dyck paths and the McCulloch machine';
         yourself
 \end{minted}
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Hi!}
+\begin{Verbatim}[fontsize=\small]
+$ whoami
+Massimo Nocentini
+PhD student @ University of Florence
+Mathematician (algebraic combinatorics, formal methods for algs)
+Programmer (automated reasoning, logics and symbolic comp)
+https://github.com/massimo-nocentini
+$ clear
+\end{Verbatim}
+I believe \textit{microkanren} is, first of all, an \textit{educational beast},
+concerning unification, lazy streams, backtracking and optimization; the
+abstract definition was shown by \textit{Dan Friedman} and \textit{Jason
+Hemann} at Scheme '13, Alexandria.
+\vfill
+I repeat the exercise of writing it in:
+\begin{itemize}
+\item Python, native \verb|generator|s :) limits on recursive calls :(
+\item OCaml, algebraic datatypes :) hard to extend :(
+\item Smalltalk, simple, fast and clear :) many dispatching msgs :/
+\end{itemize}
+
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Main idea}
+In math a relation $P$ is usually characterized by
+\begin{displaymath}
+\forall a,b,c.\,P(a,b,c) \leftrightarrow a + b = c \quad\text{entails}\quad P(1,2,3)
+\end{displaymath}
+can be expressed using either the \textit{imperative style}
+\begin{minted}[fontsize=\footnotesize]{smalltalk}
+a := 1.
+b := 2.
+c := a + b.
+Object assert: [ c = 3 ].
+\end{minted}
+or the \textit{functional style}
+\begin{minted}[fontsize=\footnotesize]{smalltalk}
+Object assert: [
+    ([ :a :b | a + b ] value: 1 value: 2) = 3 ]
+\end{minted}
+or, finally, the \textit{declarative style}
+\begin{minted}[fontsize=\footnotesize]{smalltalk}
+Object assert: [
+    [ :a :b :c | a + b = c ] value: 1 value: 2 value: 3 ]
+\end{minted}
+
 \end{frame}
@@
 \vfill
-\textit{Logical reasoning} boils down to decide the \textit{entailment}
-relation $\models$
+Let $\models$ and $\Rightarrow$ denote the \textit{entail} and \textit{imply} relations, respectively, in
 \begin{displaymath}
 \begin{split}
+\alpha &\models \beta \leftrightarrow \\
 M(\alpha) &\subseteq M(\beta) \leftrightarrow \\
-\alpha &\models \beta \leftrightarrow \\
 (\alpha &\Rightarrow \beta) \text{ is valid } \leftrightarrow\\
-\neg(\neg\alpha &\vee \beta) \text{ is unsatisfiable}
+\neg(\neg\alpha &\vee \beta) \text{ is unsatisfiable;}
 \end{split}
@@
 therefore, to prove a sentence $\alpha$ reduces to decide
-$\neg\alpha\models\perp$,\\where $\perp$ is the (\textit{unsatisfiable}) empty
-clause.
+$$\neg\alpha\models\perp\quad\leftrightarrow\quad\alpha\text{ is valid,}$$
+where $\perp$ denotes the empty clause, namely \textit{falsehood}.
 \end{frame}
@@
 \frametitle{Resolution by \textit{Refutation}}
-The \textit{resolution rule} is a \textit{complete} inference algorithm; \\
-let $l_{q}, m_{w} \in\lbrace 0,1\rbrace$ in
+The \textit{resolution rule} is a \textit{complete} inference algorithm,
 \begin{displaymath}
@@
 \end{displaymath}
-where $\left(l_{0},\ldots, l_{i}, \ldots, l_{j-1}\right) = l_{0}\vee \ldots \vee l_{i} \vee \ldots \vee l_{j-1}$.
+where $\left(l_{0},\ldots, l_{i}, \ldots, l_{j-1}\right) = l_{0}\vee \ldots
+\vee l_{i} \vee \ldots \vee l_{j-1}$, \\
+for all $l_{q}, m_{w} \in\lbrace 0,1\rbrace$.
 \vfill
@@
 $$ \lbrace y = cons(2,y) \rbrace $$
-has no \textit{finite} solution; oth, $\theta = \lbrace y \mapsto
-cons(2,cons(2,cons(2,...))) \rbrace$ is a solution by \textit{bisimulation}.
+has no \textit{finite} solution; on the other hand,
+$$\theta = \lbrace y \mapsto cons(2,cons(2,cons(2,...))) \rbrace$$
+is a solution upto \textit{bisimulation}.
 \end{frame}
@@
 \item[eliminate] if $x \not\in vars(t)$ and $x \in vars(G)$ then $$G \cup \lbrace x = t\rbrace \rightarrow G\lbrace x \mapsto t\rbrace \cup \left\lbrace x \triangleq t\right\rbrace $$
-\item[occur check] if $x \in vars(f(s_{0},\ldots,s_{k}))$ then $$G \cup \lbrace x = t(s_{0}, \ldots, s_{k}\rbrace \rightarrow \,\perp$$
+\item[occur check] if $x \in vars(f(s_{0},\ldots,s_{k}))$ then $$G \cup \lbrace x = f(s_{0}, \ldots, s_{k})\rbrace \rightarrow \,\perp$$
 \end{description}
@@
 \frametitle{microkanren}
+Let \textit{solution}, \textit{substitution} and \textit{state} be synonyms; so, $\mu$-kanren
 \begin{itemize}
-\item a \textit{DSL} for relational programming written in Scheme
-\item \textit{purely functional} implementation of \textit{miniKanren}
-\item \textit{explicit streams} of satisfying states, \textit{goal-based} approach
-\item resolution by unification via \textit{structural induction}
-\item \textit{complete}, but \textit{unfair}, search strategy
+\item is a \textit{DSL} for relational programming written in Scheme
+\item is a \textit{purely functional} core of \textit{miniKanren}
+\item provides \textit{explicit streams} of satisfying states
+\item encodes math rel using a \textit{goal-based} approach
+\item uses resolution by unification via \textit{structural induction}
 \end{itemize}
+\vfill
+A \textit{goal} is an object that responds to the \verb|#onState:| selector, it
+receives a \textit{substitution} and returns a \verb|Chain| object of
+substitutions.
+\end{frame}
 
-\iffalse
-my contribution
-- _Pythonic_ [implementation][mkpy]: functional at the core, objective at the interface
-- generators subsume _countably_-satisfiable relations; complete, _fair_ [search][dovetail]
-- _The Reasoned Schemer_ fully tested via [Travis CI][travis]; moreover, [read the docs][rtfd]
-- case studies: Smullyan puzzles and combinatorics
-- tweaking HOL Light for _certified deductions_, [wip][klight]
-\fi
+\begin{frame}[fragile]
+\frametitle{Chain hierarchy}
+We model a (possibly infinite) space of objects with the
+\verb|Chain| hierarchy, which has \verb|Bottom| and \verb|Knot|
+as subclasses which denote the empty and a populated set, respectively.
+\vfill
+Although Pharo provides the \verb|Generator| class, we write our version of
+\textit{lazy} enumeration, which is purely functional (neither clever uses of
+\verb|thisContext| nor reentrant blocks).
+\vfill
+Encodes the two \textit{monadic} operations \verb|mplus| and \verb|bind|,
+which allow us to merge two \verb|Chain|s and to combine a \verb|Chain| obj
+to yield an extended \verb|Chain| obj, respectively.
+\vfill
+Dispatch over two strategies \verb|Sequential| and \verb|Interleaved| in
+order to \textit{enumerate} solution spaces.
 \end{frame}
@@
 testNumbers
-    self assert: (self nats atMost: 10) mature equals: (0 to: 9).
+    self
+        assert: (self nats atMost: 10) mature
+        equals: (0 to: 9).
     self
@@
         equals: {10 . 5 . 16 . 8 . 4 . 2 . 1 . 4 . 2 . 1}.
-    self
-        assert: ((self gcd: 18 gcd: 32) atMost: 10) mature
-        equals: {18 . 32 . 18 . 14 . 4 . 2 . 2}
 \end{minted}
@@
 
+\iffalse
 \begin{frame}[fragile]
@@
 \end{frame}
+\fi
 
 \begin{frame}[fragile]
-\frametitle{Goals}
+\frametitle{Goal hierarchy}
+\textit{microkanren} represents math rels using the \verb|Goal| hierarchy
+\begin{itemize}
+\item \verb|Succeed| it \textit{is} satisfied by \textit{each} sub;
+\item \verb|Fail| it \textit{is not} satisfied by \textit{any} sub;
+\item \verb|Or| it is satisfied if at least one obj it consumes can be satisfied;
+\item \verb|And| it is satisfied if both objs it consumes can be satisfied;
+\item \verb|Fresh| it introduces logic vars into the goal it combines;
+\item \verb|Unify| it is satisfied if the two objs it consumes can be unified.
+\end{itemize}
+\vfill
+Moreover, a substitution (aka, a set of assignments) is represented by a
+\verb|Dictionary| obj, wrapped by \verb|State| obj to count the number of logic
+vars introduced by \verb|Fresh| goals.
+\vfill
+Our substitutions are \textit{triangular} in the sense that if
+\begin{displaymath}
+\theta =  \lbrace x \mapsto y, y \mapsto z, z \mapsto 3 \rbrace
+\end{displaymath}
+then $x \mapsto 3$ is subsumed by $\theta$,
+this is implemented in \mintinline{smalltalk}|State>>#walk|.
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{State}
 \begin{minted}[fontsize=\footnotesize]{smalltalk}
-Succeed>>onState: aState
-    ^ Chain with: aState
+State>>walk: anObj
+    | k |
+    k := anObj.
+    [ k := substitution at: k ifAbsent: [ ^ k ] ] repeat
 \end{minted}
-\begin{minted}[fontsize=\footnotesize]{smalltalk}
-Fail>>onState: aState
-    ^ Chain bottom
-\end{minted}
+A substitution is extended by
 \begin{minted}[fontsize=\footnotesize]{smalltalk}
@@
 \begin{frame}[fragile]
-\frametitle{Goals}
+\frametitle{Goal subclass: \#[Succeed | Fail | Disj | Conj]}
+In parallel, \mintinline{smalltalk}{true} and \mintinline{smalltalk}{false} have logical brothers
+\begin{minted}[fontsize=\footnotesize]{smalltalk}
+Succeed>>onState: aState
+    ^ Chain with: aState
+\end{minted}
+\begin{minted}[fontsize=\footnotesize]{smalltalk}
+Fail>>onState: aState
+    ^ Chain bottom
+\end{minted}
+respectively; btw, for conjuction and disjunction we have
+\begin{minted}[fontsize=\footnotesize]{smalltalk}
+Disj>>onState: aState
+    ^ interleaving of: ((either onState: aState)
+        mplus: [ or onState: aState ])
+\end{minted}
+\begin{minted}[fontsize=\footnotesize]{smalltalk}
+Conj>>onState: aState
+    ^ interleaving of: ((both onState: aState) bind: and)
+\end{minted}
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Goal subclass: \#Fresh}
 \begin{minted}[fontsize=\footnotesize]{smalltalk}
@@
 \begin{frame}[fragile]
-\frametitle{Goals}
+\frametitle{Goal subclass: \#Unify}
 \begin{minted}[fontsize=\footnotesize]{smalltalk}
@@
                  onState: aState
-\end{minted}
-\begin{minted}[fontsize=\footnotesize]{smalltalk}
-State>>walk: anObj
-    | k |
-    k := anObj.
-    [ k := substitution at: k ifAbsent: [ ^ k ] ] repeat
 \end{minted}
@@
 \begin{minted}[fontsize=\footnotesize]{smalltalk}
-unifyLinkedList: aCollection withLinkedList: anotherCollection onState: aState
-    ^ aCollection size = anotherCollection size
-        ifTrue: [ (aCollection zip: anotherCollection)
+unifyLinkedList: c withLinkedList: d onState: aState
+    ^ c size = d size
+        ifTrue: [ (c zip: d)
             inject: aState
-            into: [ :s :pair | self unify: pair key 
-                                    with: pair value 
-                                    onState: s ] ]
+            into: [ :s :p | self unify: p key with: p value onState: s ] ]
         ifFalse: [ UnificationError signal ]
@@
 \begin{frame}[fragile]
-\frametitle{Unifier}
-\begin{minted}[fontsize=\footnotesize]{smalltalk}
-Disj>>onState: aState
-    ^ interleaving of: ((either onState: aState) 
-        mplus: [ or onState: aState ])
-\end{minted}
-\begin{minted}[fontsize=\footnotesize]{smalltalk}
-Conj>>onState: aState
-    ^ interleaving of: ((both onState: aState) bind: and)
-\end{minted}
+\frametitle{Goal subclass: \#Cond}
 \begin{minted}[fontsize=\footnotesize]{smalltalk}
@@
 Cond>>ifPure: aStrategy
-    if := [ :clause :otherwise | 
+    if := [ :c :o |
         IfPure new
-                question: clause key 
-                    answer: clause value 
-                    otherwise: otherwise;
+            question: c key answer: c value otherwise: o;
             streamCombinationStrategy: aStrategy;
@@
 \end{minted}
-\end{frame}
-
-\begin{frame}[fragile]
-\frametitle{Unifier}
 \begin{minted}[fontsize=\footnotesize]{smalltalk}
@@
 \begin{frame}[fragile]
-\frametitle{Unifier}
+\frametitle{Dyck paths}
 Let $\mathcal{D}$ be the set of \textit{Dyck paths} and let $\leadsto$ be the \textit{CFG}
 $$
-\left\lbrace\begin{array}{l}
-\leadsto = \varepsilon \\
-\leadsto = ( \leadsto ) \leadsto \\
-\end{array}\right\rbrace
+\leadsto \,= \varepsilon \,\left|\, ( \leadsto ) \leadsto \right.
 $$
-question: enumerate $\mathcal{D}$ using $\leadsto$.
+where $\varepsilon$ is the empty string; so, \textit{enumerate} $\mathcal{D}$
+\textit{using} $\leadsto$.
 \begin{minted}[fontsize=\footnotesize]{smalltalk}
@@
 \begin{frame}[fragile]
-\frametitle{Unifier}
+\frametitle{Dyck paths}
 \begin{minted}[fontsize=\footnotesize]{smalltalk}
@@
 \begin{frame}[fragile]
-\frametitle{Unifier}
+\frametitle{McCulloch's machine and the MC lock puzzle}
 Let $X$ and $Y$ be natural numbers in machine
@@
 \end{displaymath}
-question: does exist a number $\alpha$ such that $ \alpha \stackrel{\circ}{\rightarrow} \alpha $?
+question: \textit{does exist a number} $\alpha$ \textit{such that} $ \alpha \stackrel{\circ}{\rightarrow} \alpha $?
 \begin{minted}[fontsize=\footnotesize]{smalltalk}
@@
 \begin{frame}[fragile]
-\frametitle{Unifier}
+\frametitle{McCulloch's machine and the MC lock puzzle}
 \begin{minted}[fontsize=\footnotesize]{smalltalk}
@@
 
+\begin{frame}[fragile]
+\frametitle{End}
+A quick check:
+\\\\
+\begin{overprint}
+    \onslide<1>
+    \begin{tabular}{c}
+        $\displaystyle { \over 323 \stackrel{\circ}{\rightarrow}\quad } $\\
+    \end{tabular}
+    \onslide<2>
+    \begin{tabular}{c}
+        $\displaystyle { \over 23 \stackrel{\circ}{\rightarrow}\quad } $\\
+        $\displaystyle {   \over 323 \stackrel{\circ}{\rightarrow}\quad } $\\
+    \end{tabular}
+    \onslide<3>
+    \begin{tabular}{c}
+        $\displaystyle { \over 23 \stackrel{\circ}{\rightarrow} 3} $\\
+        $\displaystyle {   \over 323 \stackrel{\circ}{\rightarrow}\quad } $\\
+    \end{tabular}
+    \onslide<4>
+    \begin{tabular}{c}
+        $\displaystyle { \over 23 \stackrel{\circ}{\rightarrow} 3} $\\
+        $\displaystyle {   \over 323 \stackrel{\circ}{\rightarrow}323 } $\\
+    \end{tabular}
+    \onslide<5>
+    \begin{tabular}{c}
+        $\displaystyle { \over 23 \stackrel{\circ}{\rightarrow} 3} $\\
+        $\displaystyle {   \over 323 \stackrel{\circ}{\rightarrow}323 } $\\
+    \end{tabular}
+    \\\\\\\\
+    Future directions:
+    \begin{itemize}
+        \item unification is just a \textit{constraint}...
+        \item ...so add \textit{disequality}, \textit{type checks} and other constraints
+        \item \textit{impure} operators, such as Prolog's \textit{cut} (!)
+        \item automatic message dispatching for unifications
+    \end{itemize}
+\end{overprint}
+
+\end{frame}
+
+\begin{frame}{ }
+\Huge Thanks!
+\end{frame}
+
 \end{document}

### 2018-12-05 booklet/article.tex
Add initial content to the booklet.

\documentclass[a4paper,12pt]{article}

\usepackage{inputenc}
\usepackage{euler}
\usepackage[T1]{fontenc}
\usepackage{libertine}
\usepackage{lipsum}
\usepackage{fancyvrb}
\usepackage{url}
\usepackage[english]{babel}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{acronym}
\usepackage{hyperref}
\usepackage{tabu}
\usepackage{rotating}
\usepackage{mathdots}
\usepackage{minted}
\usepackage{units}
\usepackage{float}
\usepackage{bbold}
\usepackage[sort&compress,square,comma,authoryear]{natbib}

\fvset{fontsize=\normalsize}
\setmonofont[Scale=0.8]{Menlo}
\usemintedstyle{xcode}
\hypersetup{colorlinks=false, citecolor=Violet}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{example}[theorem]{Example}

\author{Massimo Nocentini}
\title{Relational programming in Smalltalk}

\begin{document}

\maketitle

\begin{abstract}
This booklet is an extended description of $mu$Kanren, a relational interpreter
formerly implemented in Scheme and ported in Smalltalk for the sake of
understanding and, of course, fun. It enjoy simplicity and elegance despite the
complex nature of logic systems; moreover, using an uniform but powerful
language such as Smalltalk it is possible to appreciate and benefit from them
all.
\end{abstract}

\section{Introduction}

I believe \textit{microkanren} is, first of all, an \textit{educational beast},
concerning unification, lazy streams, backtracking and optimization; the
abstract definition was shown in \citep{Hemann:muKanren}.
I repeat the exercise of writing it in \citep{Nocentini:kanrens}:
\begin{itemize}
\item Python, native \verb|generator|s :) limits on recursive calls :(
\item OCaml, algebraic datatypes :) hard to extend :(
\item Smalltalk, simple, fast and clear :) many dispatching msgs :/
\end{itemize}

\begin{example}
In math a relation $P$ is usually characterized by
\begin{displaymath}
\forall a,b,c.\,P(a,b,c) \leftrightarrow a + b = c \quad\text{entails}\quad P(1,2,3)
\end{displaymath}
can be expressed using either the \textit{imperative style}
\begin{minted}[fontsize=\footnotesize]{smalltalk}
a := 1.  b := 2.  c := a + b.  Object assert: [ c = 3 ]
\end{minted}
or the \textit{functional style}
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Object assert: [ ([ :a :b | a + b ] value: 1 value: 2) = 3 ]
\end{minted}
or, finally, the \textit{declarative style}
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Object assert: [ [ :a :b :c | a + b = c ] value: 1 value: 2 value: 3 ]
\end{minted}
which closely resembles the mathematical statement.
\end{example}

We quickly review some mathematical backgrounds underlying logic systems;
in particular deductions are performed according to \emph{resolutions}

\begin{description}

\item[by refutation] Let $\alpha$ be a sentence in \emph{conjunctive normal
form} (CNF for short) and $M(\alpha)$ the set of models that satisfy $\alpha$
-- recall that a \emph{model} is a set of assignments of values to free
variables in $\alpha$ that make the whole sentence true.  We say that, on one
hand, $\alpha$ is \textit{valid} if it is true in \textit{all} models; on the
other hand, $\alpha$ is \textit{satisfiable} if it is true in \textit{some}
model.

Let $\models$ and $\Rightarrow$ denote the \textit{entail} and \textit{imply}
relations, respectively; so,
\begin{displaymath}
\begin{split}
\alpha \models \beta \leftrightarrow
M(\alpha) \subseteq M(\beta) \leftrightarrow
(\alpha \Rightarrow \beta) \text{ is valid } \leftrightarrow
\neg(\neg\alpha \vee \beta) \text{ is unsatisfiable.}
\end{split}
\end{displaymath}
Therefore, to prove a sentence $\alpha$ reduces to prove
$\neg\alpha\models\perp$, where $\perp$ denotes the empty clause $()$, namely
\textit{falsehood}, which can be deduced by the \textit{resolution rule}
introduced in \citep{Robinson:1965:MLB:321250.321253}; if $l_{i}$ and $m_{r}$
bound to the same variable then
\begin{displaymath}
{\left(l_{0}, \ldots, l_{i}, \ldots, l_{j-1}\right) \quad \left(m_{0}, \ldots, m_{r},\ldots, m_{k-1}\right) \quad l_{i} = \neg m_{r}
\over
\left(l_{0},\ldots, l_{i-1},l_{i+1}, \ldots,l_{j-1}, m_{0},\ldots, m_{r-1},m_{r+1},\ldots, m_{k-1}\right)}
\end{displaymath}
where $\left(l_{0},\ldots, l_{i}, \ldots, l_{j-1}\right) = l_{0}\vee \ldots
\vee l_{i} \vee \ldots \vee l_{j-1}$, for all $l_{q}, m_{w} \in\lbrace 0,1\rbrace$.
This rule is a \textit{complete} inference algorithm, namely it enumerates
\emph{all} deductions for a given sentence, although requiring
\emph{exponential} time; for the sake of concreteness, the \textit{DPLL}
algorithm \citep{Davis:1962:MPT:368273.368557}
is a recursive, depth-first enumeration of models using the
resolution rule paired with heuristics \textit{early termination},
\textit{pure symbol} and \textit{unit clause} to speed up.

\item[by unification] is a the process introduced in \citep{robinson_unif} and
improved by \citep{Martelli:1982:EUA:357162.357169}; it consists of solving
\textit{equations among symbolic expressions}. A \textit{solution} is denoted
as a \textit{substitution} $\theta$, namely a mapping that assigns a symbolic
values to free variables; let $G$ be a set of equations, unification proceeds
according to the following rules:
\begin{description}
\item[delete] $G \cup \lbrace t = t \rbrace \rightarrow G$
\item[decompose] $G \cup \lbrace f(s_{0}, \ldots, s_{k}) = f(t_{0}, \ldots, t_{k})\rbrace$ entails
$$G \cup \lbrace s_{0}=t_{0},\ldots, s_{k}=t_{k} \rbrace$$
\item[conflict] if $f\neq g \vee k\neq m$ then $$G \cup \lbrace f(s_{0}, \ldots, s_{k}) = g(t_{0}, \ldots, t_{m})\rbrace \rightarrow \,\perp$$
\item[eliminate] if $x \not\in vars(t)$ and $x \in vars(G)$ then $$G \cup \lbrace x = t\rbrace \rightarrow G\lbrace x \mapsto t\rbrace \cup \left\lbrace x \triangleq t\right\rbrace $$
\item[occur check] if $x \in vars(f(s_{0},\ldots,s_{k}))$ then $$G \cup \lbrace x = f(s_{0}, \ldots, s_{k})\rbrace \rightarrow \,\perp;$$
without it, generating a $\theta$ is a
\emph{recursive enumerable} problem.
\end{description}

\end{description}

\begin{example}
Let $x$ and $y$ be free variables, the set
$$\lbrace cons(x,cons(x,nil)) = cons(2,y)\rbrace$$
has solution $\theta = \lbrace x \mapsto 2, y \mapsto cons(2,nil) \rbrace$;
on the contrary, the set
$$ \lbrace y = cons(2,y) \rbrace $$
has no \textit{finite} solution, nonetheless
$$\theta = \lbrace y \mapsto cons(2,cons(2,cons(2,...))) \rbrace$$
is a solution upto \textit{bisimulation}
\citep{10.1007/BFb0017309, DBLP:books/daglib/0067019}.
\end{example}

\section{Conclusions}

A quine generator is described in \citep{Byrd:2012:MLU:2661103.2661105}.

\bibliographystyle{plainnat}
\bibliography{biblio}
\end{document}

### 2018-12-07 booklet/article.tex
Little more introduction paragraphs.

@@
 \begin{abstract}
-This booklet is an extended description of $mu$Kanren, a relational interpreter
-formerly implemented in Scheme and ported in Smalltalk for the sake of
-understanding and, of course, fun. It enjoy simplicity and elegance despite the
-complex nature of logic systems; moreover, using an uniform but powerful
-language such as Smalltalk it is possible to appreciate and benefit from them
-all.
+This booklet is an extended description of $\mu$Kanren, a relational
+interpreter formerly implemented in Scheme and ported in Smalltalk for the sake
+of understanding and, of course, fun. It enjoy simplicity and elegance despite
+the complex nature of logic systems; moreover, using an uniform but powerful
+language such as Smalltalk it is possible to appreciate and benefit from both
+of them.
 \end{abstract}
@@
 
+This document extends the presentation of $\mu$Kanren shown at the last ESUG
+conference \citep{ESUG2018} in two ways: (i)~it recaps existing ``literature''
+on the subject and (ii)~it describes in more details our implementation coded
+in Smalltalk.
+
+It is a waste of time to repeat ourself on $\mu$Kanren's definition; on the
+contrary, we believe interesting to look up and to watch over the many
+implementation it has inspired since its seminal paper \citep{Hemann:muKanren}
+and how programmers stretch the canonical implementation to take advantage of
+their favorite programming language's features.
+
+Certainly the book \citep{Friedman:Reasoned:Schemer} is the main reference
+written in the same style of the other \emph{little books} of which Dan
+Friedman coauthors them all; moreover, Byrd's dissertation \citep{Byrd:PhD}
+shows advanced stuff and techniques while the repository
+\citep{byrd:minikanren:repo} -- updated regularly by Byrd himself -- keeps
+pointers to implementations, articles and presentations. All these references
+uses the Scheme programming language, $\mu$Kanren is \emph{purely functional}
+therefore any language that has the same power of the $\lambda$-calculus is a
+candidate for implementing it, Scheme in particular; in parallel,
+\citep{byrd:minikanren} has the same nature but uses Scheme macrology and in
+some aspect it requires little effort to port to other languages that haven't
+macros.
+
 I believe \textit{microkanren} is, first of all, an \textit{educational beast},
 concerning unification, lazy streams, backtracking and optimization; the
-abstract definition was shown in \citep{Hemann:muKanren}.
+abstract definition was shown in 
 I repeat the exercise of writing it in \citep{Nocentini:kanrens}:

### 2018-12-12 booklet/article.tex
Add more comments on references.

@@
 
-I believe \textit{microkanren} is, first of all, an \textit{educational beast},
-concerning unification, lazy streams, backtracking and optimization; the
-abstract definition was shown in 
-I repeat the exercise of writing it in \citep{Nocentini:kanrens}:
+Another implementation using the \emph{Clojure} dialect of Lisp is
+\citep{clojure:core.logic} which targets the JVM an provides constraints and
+nominal logic programming; on the other hand, \citep{sullivan:microkanrenhs}
+is a clean and concise implementation using Haskell, while many others simpler
+versions can be found, written for the sake of understanding and, ultimately,
+for fun.
+
+Finally, your humble authors believe that $\mu$Kanren is a sophisticated
+\textit{educational machine} because it concerns and mix together unification,
+lazy streams, backtracking and optimization techniques; moreover, it allows
+programmers to write their own version according to their personal taste and
+style, ranging from implementations pretty imperative to ones built over
+categories.  We repeat the exercise of writing it in \citep{Nocentini:kanrens}
+targeting different languages, in particular
 \begin{itemize}
-\item Python, native \verb|generator|s :) limits on recursive calls :(
-\item OCaml, algebraic datatypes :) hard to extend :(
-\item Smalltalk, simple, fast and clear :) many dispatching msgs :/
+\item using \emph{Python} we take advantage of native \verb|generator|s objects
+and we provide a \emph{fair} enumeration strategy that the canonical one
+hasn't; on the contrary, we experience computational limits because of
+\texttt{RecursionError} exceptions. This implementation is complete and
+parallels the one shown in the reference book, accompanied by a unit test suite
+that covers \emph{all} questions explained in the book and some new
+applications such as (i)~finding fixed points of a set of inference rules and
+enumeration of classes of combinatorial objects.
+\item using \emph{OCaml} the syntax of the language can be cleanly defined by
+algebraic datatypes and the type system helps in the design of the required
+operators.  While object-oriented languages make extensibility a relative
+simple task, this implementation is harder unless we can make changes to basic
+types, at least for what concerns the central \emph{unification} procedure.
+All of this is preparatory to modify the tactics mechanism of the \emph{HOL
+Light} theorem prover in order to introduce the relational paradigm in that
+environment.
 \end{itemize}
+The current work describe a Smalltalk implementation of the basic system
+with the aim to learn and experiment with a pure object environment, such as 
+Pharo and Squeak. Starting small, the next example we show three different
+kinds to perform an identity among \verb|SmallInteger| objects:
 
 \begin{example}
-In math a relation $P$ is usually characterized by
+In math a relation $P$ is usually characterized by a statement such as
 \begin{displaymath}
-\forall a,b,c.\,P(a,b,c) \leftrightarrow a + b = c \quad\text{entails}\quad P(1,2,3)
+\forall a,b,c.\,P(a,b,c) \leftrightarrow a + b = c \quad\text{from which}\quad
+P(1,2,3) \quad\text{can be proved.}
 \end{displaymath}
-can be expressed using either the \textit{imperative style}
+The same check can be expressed in Smalltalk using either the
+\textit{imperative style}
 \begin{minted}[fontsize=\footnotesize]{smalltalk}
@@
 \end{example}
+
+\section{Resolution methods}
 

### 2019-03-06 booklet/article.tex
Description of `Succeed` and `Fail` goal for the booklet.

@@
 
+\newcommand{\stcode}[1]{\begin{minted}[fontsize=\footnotesize]{smalltalk} #1 \end{minted}}
+
 \author{Massimo Nocentini}
@@
 
-\section{Introduction}
+\section*{Introduction}
 
@@
 
-\section{Resolution methods}
+\tableofcontents
+
+\section{Bootstrapping $\mu$Kanren}
+
+We believe that an unusual development process can be done in Smalltalk and it
+is the one that had driven and currently drives the present description.
+Instead of creating classes first, we start by looking at some \textit{existing
+object} that can be a good candidate to take into account our messages; in
+sight of this approach, we give the fundamentals of $\mu$Kanren -- more
+generally, for a class of logic programming languages -- starting from the two
+nice boolean objects.
+
+\subsection{\texttt{Succeed} and \texttt{Fail} classes}
+
+Let us consider the object \Verb|false| whose responsibility is to represent
+\textit{falsehood}, hence it should be a good candidate to provide us something
+related to this from the \textit{relational} point of view,
+\begin{minted}[fontsize=\footnotesize]{smalltalk}
+false asGoal
+\end{minted}
+namely, \Verb|false| can build for us an object that represents a relation that
+cannot be satisfied at all responding to the message selector \Verb|asGoal| --
+don't be scared by multiple names, we freely interchange the terms
+\textit{goal} and \textit{relation} postponing to later sections the
+description of their technical difference. Moreover, let us 
+assume that the infrastructure that actually build results is given and implemented
+by the \textit{oracle block} \texttt{B := [:g | g solutions all]}, we write our first test case:
+\begin{minted}[fontsize=\footnotesize]{smalltalk}
+GoalTest>>#testFail
+    self 
+        assert: false asGoal solutions all 
+        equals: {}
+\end{minted}
+Now we get within the \Verb|False| class to add the implementation of the message
+\begin{minted}[fontsize=\footnotesize]{smalltalk}
+False>>#asGoal
+    ^ Goal fail
+\end{minted}
+and of the message
+\begin{minted}[fontsize=\footnotesize]{smalltalk}
+Goal>>#fail
+    ^ Fail new
+\end{minted}
+such that to say that \verb|Fail| denotes the impossibility to satisfy a relation
+is the same to code the message
+\begin{minted}[fontsize=\footnotesize]{smalltalk}
+Fail>>#onState: aState 
+    ^ Chain bottom
+\end{minted}
+which will be used within the message \Verb|solutions| sent in the oracle block
+\Verb|B| -- a lot remains to be explained, such as the \Verb|Chain| class and
+streams in particular, all in what follows.
+
+In parallel, we reproduce for the object \Verb|true| the steps did for
+the object \Verb|false|; first,
+\begin{minted}[fontsize=\footnotesize]{smalltalk}
+GoalTest>>#testSucceed
+    self 
+        assert: true asGoal solutions all 
+        equals: {Var tautology}
+\end{minted}
+second,
+\begin{minted}[fontsize=\footnotesize]{smalltalk}
+True>>#asGoal
+    ^ Goal succeed
+\end{minted}
+third,
+\begin{minted}[fontsize=\footnotesize]{smalltalk}
+Goal>>#succeed
+    ^ Succeed new
+\end{minted}
+fourth,
+\begin{minted}[fontsize=\footnotesize]{smalltalk}
+Succeed>>#onState: aState 
+    ^ Chain with: aState
+\end{minted}
+
+\section{Theoretical backgrounds}
 
@@
 
-\begin{description}
+\subsection{Resolution by refutation}
 
-\item[by refutation] Let $\alpha$ be a sentence in \emph{conjunctive normal
+Let $\alpha$ be a sentence in \emph{conjunctive normal
 form} (CNF for short) and $M(\alpha)$ the set of models that satisfy $\alpha$
@@
 
-\item[by unification] is a the process introduced in \citep{robinson_unif} and
+\subsection{Resolution by unification}
+
+This method was introduced in \citep{robinson_unif} and
 improved by \citep{Martelli:1982:EUA:357162.357169}; it consists of solving
@@
 
-\end{description}
-
 \begin{example}
@@
 \end{example}
-
 \section{Conclusions}
 
-A quine generator is described in \citep{Byrd:2012:MLU:2661103.2661105}.
+A quine generator is described in \citep{Byrd:2012:MLU:2661103.2661105} and
+we aim to reproduce it.
 

### 2019-03-07 booklet/article.tex
Apply fixes by Stef Ducasse.

@@
 interpreter formerly implemented in Scheme and ported in Smalltalk for the sake
-of understanding and, of course, fun. It enjoy simplicity and elegance despite
+of understanding and, of course, fun. It enjoys simplicity and elegance despite
 the complex nature of logic systems; moreover, using an uniform but powerful
@@
 contrary, we believe interesting to look up and to watch over the many
-implementation it has inspired since its seminal paper \citep{Hemann:muKanren}
+implementations it has inspired since its seminal paper \citep{Hemann:muKanren}
 and how programmers stretch the canonical implementation to take advantage of
@@
 Friedman coauthors them all; moreover, Byrd's dissertation \citep{Byrd:PhD}
-shows advanced stuff and techniques while the repository
-\citep{byrd:minikanren:repo} -- updated regularly by Byrd himself -- keeps
+shows advanced stuffs and techniques while the repository
+\url{http://minikanren.org/} -- updated regularly by Byrd himself -- keeps
 pointers to implementations, articles and presentations. All these references
-uses the Scheme programming language, $\mu$Kanren is \emph{purely functional}
+use the Scheme programming language, $\mu$Kanren is \emph{purely functional}
 therefore any language that has the same power of the $\lambda$-calculus is a
-candidate for implementing it, Scheme in particular; in parallel,
-\citep{byrd:minikanren} has the same nature but uses Scheme macrology and in
-some aspect it requires little effort to port to other languages that haven't
-macros.
+candidate for implementing it, Scheme in particular; in parallel, the
+implementation with symbolic constraints has the same nature but uses Scheme
+macrology and in some aspect it requires little effort to port to other
+languages that haven't macros.
 
 Another implementation using the \emph{Clojure} dialect of Lisp is
-\citep{clojure:core.logic} which targets the JVM an provides constraints and
+\citep{clojure:core.logic} which targets the JVM and provides constraints and
 nominal logic programming; on the other hand, \citep{sullivan:microkanrenhs}
@@
 Finally, your humble authors believe that $\mu$Kanren is a sophisticated
-\textit{educational machine} because it concerns and mix together unification,
+\textit{educational machine} because it concerns and mixes together unification,
 lazy streams, backtracking and optimization techniques; moreover, it allows
@@
 \end{itemize}
-The current work describe a Smalltalk implementation of the basic system
+The current work describes a Smalltalk implementation of the basic system
 with the aim to learn and experiment with a pure object environment, such as 
 Pharo and Squeak. Starting small, the next example we show three different
-kinds to perform an identity among \verb|SmallInteger| objects:
+ways to express equality between \verb|SmallInteger| objects:
 
@@
 description of their technical difference. Moreover, let us 
-assume that the infrastructure that actually build results is given and implemented
-by the \textit{oracle block} \texttt{B := [:g | g solutions all]}, we write our first test case:
+assume that the infrastructure that actually builds results is given and implemented
+by the block \texttt{B := [:g | g solutions all]}, we write our first test case:
 \begin{minted}[fontsize=\footnotesize]{smalltalk}
@@
 \end{minted}
-Now we get within the \Verb|False| class to add the implementation of the message
+Now we extend the \Verb|False| class with the following \Verb|asGoal| message
+definition
 \begin{minted}[fontsize=\footnotesize]{smalltalk}
@@
 \end{minted}
-which will be used within the message \Verb|solutions| sent in the oracle block
+which will be used within the message \Verb|solutions| sent in the block
 \Verb|B| -- a lot remains to be explained, such as the \Verb|Chain| class and
@@
 We quickly review some mathematical backgrounds underlying logic systems;
-in particular deductions are performed according to \emph{resolutions}
+in particular deductions are performed according to \emph{resolution} methods.
 

### 2019-03-13 booklet/article.tex
The selector Object>>#cons: now double dispatches to gain more flexibility.

@@
 
-\newcommand{\stcode}[1]{\begin{minted}[fontsize=\footnotesize]{smalltalk} #1 \end{minted}}
+\newcommand{\stcode}[1]{\mintinline[fontsize=\footnotesize]{smalltalk}{#1}}
 
@@
 language such as Smalltalk it is possible to appreciate and benefit from both
-of them.
+properties.
 \end{abstract}
@@
 
-This document extends the presentation of $\mu$Kanren shown at the last ESUG
-conference \citep{ESUG2018} in two ways: (i)~it recaps existing ``literature''
-on the subject and (ii)~it describes in more details our implementation coded
-in Smalltalk.
-
-It is a waste of time to repeat ourself on $\mu$Kanren's definition; on the
-contrary, we believe interesting to look up and to watch over the many
-implementations it has inspired since its seminal paper \citep{Hemann:muKanren}
-and how programmers stretch the canonical implementation to take advantage of
-their favorite programming language's features.
-
-Certainly the book \citep{Friedman:Reasoned:Schemer} is the main reference
-written in the same style of the other \emph{little books} of which Dan
-Friedman coauthors them all; moreover, Byrd's dissertation \citep{Byrd:PhD}
-shows advanced stuffs and techniques while the repository
-\url{http://minikanren.org/} -- updated regularly by Byrd himself -- keeps
-pointers to implementations, articles and presentations. All these references
-use the Scheme programming language, $\mu$Kanren is \emph{purely functional}
-therefore any language that has the same power of the $\lambda$-calculus is a
-candidate for implementing it, Scheme in particular; in parallel, the
-implementation with symbolic constraints has the same nature but uses Scheme
-macrology and in some aspect it requires little effort to port to other
-languages that haven't macros.
-
-Another implementation using the \emph{Clojure} dialect of Lisp is
-\citep{clojure:core.logic} which targets the JVM and provides constraints and
-nominal logic programming; on the other hand, \citep{sullivan:microkanrenhs}
-is a clean and concise implementation using Haskell, while many others simpler
-versions can be found, written for the sake of understanding and, ultimately,
-for fun.
-
-Finally, your humble authors believe that $\mu$Kanren is a sophisticated
-\textit{educational machine} because it concerns and mixes together unification,
-lazy streams, backtracking and optimization techniques; moreover, it allows
-programmers to write their own version according to their personal taste and
-style, ranging from implementations pretty imperative to ones built over
-categories.  We repeat the exercise of writing it in \citep{Nocentini:kanrens}
-targeting different languages, in particular
-\begin{itemize}
-\item using \emph{Python} we take advantage of native \verb|generator|s objects
-and we provide a \emph{fair} enumeration strategy that the canonical one
-hasn't; on the contrary, we experience computational limits because of
-\texttt{RecursionError} exceptions. This implementation is complete and
-parallels the one shown in the reference book, accompanied by a unit test suite
-that covers \emph{all} questions explained in the book and some new
-applications such as (i)~finding fixed points of a set of inference rules and
-enumeration of classes of combinatorial objects.
-\item using \emph{OCaml} the syntax of the language can be cleanly defined by
-algebraic datatypes and the type system helps in the design of the required
-operators.  While object-oriented languages make extensibility a relative
-simple task, this implementation is harder unless we can make changes to basic
-types, at least for what concerns the central \emph{unification} procedure.
-All of this is preparatory to modify the tactics mechanism of the \emph{HOL
-Light} theorem prover in order to introduce the relational paradigm in that
-environment.
-\end{itemize}
-The current work describes a Smalltalk implementation of the basic system
-with the aim to learn and experiment with a pure object environment, such as 
-Pharo and Squeak. Starting small, the next example we show three different
-ways to express equality between \verb|SmallInteger| objects:
-
-\begin{example}
-In math a relation $P$ is usually characterized by a statement such as
-\begin{displaymath}
-\forall a,b,c.\,P(a,b,c) \leftrightarrow a + b = c \quad\text{from which}\quad
-P(1,2,3) \quad\text{can be proved.}
-\end{displaymath}
-The same check can be expressed in Smalltalk using either the
-\textit{imperative style}
-\begin{minted}[fontsize=\footnotesize]{smalltalk}
-a := 1.  b := 2.  c := a + b.  Object assert: [ c = 3 ]
-\end{minted}
-or the \textit{functional style}
-\begin{minted}[fontsize=\footnotesize]{smalltalk}
-Object assert: [ ([ :a :b | a + b ] value: 1 value: 2) = 3 ]
-\end{minted}
-or, finally, the \textit{declarative style}
-\begin{minted}[fontsize=\footnotesize]{smalltalk}
-Object assert: [ [ :a :b :c | a + b = c ] value: 1 value: 2 value: 3 ]
-\end{minted}
-which closely resembles the mathematical statement.
-\end{example}
+\input{introduction.tex}
 
 \tableofcontents
+
+\section{Preparing the wires...}
+
+\subsection{Goals and substitutions}
+
+A \textit{goal} is any object that responds to the selector \Verb|#onState:| so
+that it receives a substitution and produces an ordered sequence of zero or
+more substitutions. In turn, a substitution is an unordered collection of
+values which are indexed by \textit{logic variables}, therefore a
+\Verb|Dictionary| object fits perfectly; we wrap it in a \Verb|State| object
+that also keeps record of an incremental numbering \Verb|birthdate| whose
+meaning will be clarified later when we'll describe the \Verb|Fresh|
+goal\footnote{put here the reference to its subsection}. An empty substitution
+is created by \stcode{Dictionary new asState} where
+\begin{minted}[fontsize=\footnotesize]{smalltalk}
+Dictionary>>#asState
+    ^ State new 
+        birthdate: 0;
+        substitution: self;
+        yourself
+\end{minted}
+and it can be augmented by
+\begin{minted}[fontsize=\footnotesize]{smalltalk}
+State>>#at: aVar put: aValue 
+    | s |
+    s := substitution copy.
+    s "check that if a key already exists at aVar, then the values must be the same"
+      at: aVar
+      ifPresent: [:v | aValue = v ifFalse: [UnificationError signal]]
+      ifAbsentPut: aValue.
+    ^ self class new 
+        birthdate: birthdate;
+        substitution: s;
+        yourself
+\end{minted}
+mantaining a protocol coherent to the one of \Verb|Dictionary| objects for
+extension.
+
+\subsection{Chains}
+
+Because the sequence of substitutions may be infinite, we
+represent it not as a \Verb|LinkedList| object but as a \Verb|Chain| object;
+as usual, we find \Verb|BlockClosure| objects good candidates to represent (a possibly
+infinite) computation. Some examples are in order,
+\begin{minted}[fontsize=\footnotesize]{smalltalk}
+ChainTest>>#fibs
+    ^ self fib: 0 fib: 1
+
+ChainTest>>#fib: m fib: n 
+    ^ [self fib: n fib: m + n] links: m
+
+ChainTest>>#gcd: m gcd: n 
+    ^ [n isZero 
+            ifTrue: [Chain with: m]
+            ifFalse: [self gcd: n gcd: m \\ n]] 
+        links: m
+
+ChainTest>>#nats
+    ^ self ints: 0
+
+ChainTest>>#ints: i 
+    ^ [self ints: i + 1] links: i
+
+ChainTest>>#upto2
+    ^ [[[Chain bottom] links: 2] links: 1] links: 0
+\end{minted}
+where blocks define the message \Verb|#links:| as 
+\begin{minted}[fontsize=\footnotesize]{smalltalk}
+BlockClosure>>#links: anObj
+    ^ Chain item: anObj linker: self
+\end{minted}
+exposing the abstract class \Verb|Chain| and its class-side messages
+\begin{minted}[fontsize=\footnotesize]{smalltalk}
+Chain class>>#bottom
+    ^ Bottom new
+
+Chain class>>#item: anObj linker: aBlockClosure 
+    ^ Knot new 
+        item: anObj;
+        linker: aBlockClosure;
+        yourself
+
+Chain class>>#with: anObj 
+    ^ [Chain bottom] links: anObj
+
+Chain class>>#repeat: anObj 
+    ^ [self repeat: anObj] links: anObj
+\end{minted}
+together with its two subclasses \Verb|Knot| and \Verb|Bottom|; a comprehensive
+test case follows
+\begin{minted}[fontsize=\footnotesize]{smalltalk}
+ChainTest>>#testNumbers
+    self 
+        assert: Chain bottom contents 
+        equals: LinkedList new.
+    self
+        assert: (self nats next: 10) contents
+        equals: (0 to: 9).
+    self
+        assert: (self upto2 next: 10) contents
+        equals: (0 to: 2).
+    self
+        assert: (self upto2 next: 1) contents
+        equals: (0 to: 0).
+    self 
+        assert: (self fibs next: 10) contents 
+        equals: {0. 1. 1. 2. 3. 5. 8. 13. 21. 34}.
+    self 
+        assert: ((self collatz: 10) next: 10) contents 
+        equals: {10. 5. 16. 8. 4. 2. 1. 4. 2. 1}.
+    self 
+        assert: ((self gcd: 18 gcd: 32) next: 10) contents 
+        equals: {18. 32. 18. 14. 4. 2. 2}.
+    self 
+        assert: ((Chain repeat: 4) next: 10) contents 
+        equals: {4. 4. 4. 4. 4. 4. 4. 4. 4. 4}
+\end{minted}
+On one hand, when a \Verb|Chain| object (i)~receives the selector
+\Verb|#next:| that carries a \Verb|SmallInteger| object $i$ it keeps at most
+$i$ objects and then truncate itself, precisely  
+\begin{minted}[fontsize=\footnotesize]{smalltalk}
+Bottom>>#next: anInteger 
+    ^ self
+
+Knot>>#next: n 
+    ^ n isZero
+        ifTrue: [Chain bottom]
+        ifFalse: [[self next value next: n - 1] links: item]
+\end{minted}
+On the other hand, when a \Verb|Chain| object (ii)~receives the selector
+\Verb|#contents| it builds a \Verb|LinkedList| object \textit{eagerly},
+\begin{minted}[fontsize=\footnotesize]{smalltalk}
+Bottom>>#contents
+    ^ LinkedList new
+
+Knot>>#contents
+    ^ self next contents 
+        addFirst: item;
+        yourself
+\end{minted}
+where \Verb|Knot| objects provide the following simple but fundamental message
+\begin{minted}[fontsize=\footnotesize]{smalltalk}
+Knot>>#next
+    ^ linker value
+\end{minted}
+Observe that the \Verb|Chain| hierarchy support part of the protocol provided
+by\Verb|ReadStream| objects because of the messages \Verb|#next|, \Verb|#next:|
+and \Verb|#contents|.
 
@@
 \end{example}
-\section{Conclusions}
+\section{Conclusions and further works}
 
-A quine generator is described in \citep{Byrd:2012:MLU:2661103.2661105} and
-we aim to reproduce it.
+\begin{itemize}
+\item A quine generator is described in \citep{Byrd:2012:MLU:2661103.2661105}
+and we aim to reproduce it.
+
+\item using \emph{OCaml} the syntax of the language can be cleanly defined by
+algebraic datatypes and the type system helps in the design of the required
+operators.  While object-oriented languages make extensibility a relative
+simple task, this implementation is harder unless we can make changes to basic
+types, at least for what concerns the central \emph{unification} procedure.
+All of this is preparatory to modify the tactics mechanism of the \emph{HOL
+Light} theorem prover in order to introduce the relational paradigm in that
+environment.
+\end{itemize}
 

### 2019-03-13 booklet/introduction.tex
Split the introduction and comment the Chain hierarchy.

This document extends the presentation of $\mu$Kanren shown at the last ESUG
conference \citep{ESUG2018} in two ways: (i)~it recaps existing literature 
on the subject and (ii)~it describes in more details our implementation coded
in Smalltalk.

It is a waste of time to repeat ourself on $\mu$Kanren's definition; on the
contrary, we believe interesting to look up and to watch over the many
implementations it has inspired since its seminal paper \citep{Hemann:muKanren}
and how programmers stretch the canonical implementation to take advantage of
their favorite programming language's features.

Certainly the book \citep{Friedman:Reasoned:Schemer} is the main reference
written in the same style of the other \emph{little books} of which Dan
Friedman coauthors them all; moreover, Byrd's dissertation \citep{Byrd:PhD}
shows advanced stuffs and techniques while the repository
\url{http://minikanren.org/} -- updated regularly by Byrd himself -- keeps
pointers to implementations, articles and presentations.  $\mu$Kanren is
\emph{purely functional} therefore any language that has the same power of the
$\lambda$-calculus is a candidate for implementing it, nonetheless its first
definition has been written in Scheme; in parallel, the implementation with
symbolic constraints has the same nature but uses macros and in some aspect it
requires more effort to port it to other languages that haven't macros.

Another implementation using the \emph{Clojure} dialect of Lisp is
\citep{clojure:core.logic} which targets the JVM and provides constraints and
nominal logic programming; on the other hand, \citep{sullivan:microkanrenhs}
is a clean and concise implementation using Haskell, while many others simpler
versions can be found, written for the sake of understanding and, ultimately,
for fun.

Finally, your humble authors believe that $\mu$Kanren is a sophisticated
\textit{educational machine} because it concerns and mixes together unification,
lazy streams, backtracking and optimization techniques; moreover, it allows
programmers to write their own version according to their personal taste and
style, ranging from implementations pretty imperative to ones built over
categories.

Our effort to understand $\mu$Kanren started in \citep{Nocentini:kanrens} using
the \emph{Python} programming language; we take advantage of native
\Verb|generator|s objects and we provide a \emph{fair} enumeration strategy
that the original implementation hasn't; on the contrary, we experience
computational limits because of \texttt{RecursionError} exceptions. This
implementation is complete and parallels the one shown in the reference book,
accompanied by a unit test suite that provides answers to \emph{all} questions
proposed in the book and some new applications such as (i)~finding fixed points
of a set of inference rules and (ii)~enumerating classes of combinatorial
objects. Moreover, two \textit{impure} operators, as the \textit{cut}
(\Verb|!|) Prolog operator, had been coded too. However complete, the code is a
mixture of the object orientation at the surface and functional at the core both
supported by primitives of the underlying language, such as \Verb|yield|
expressions and magic methods; took note of this, we desire something more
uniform and cohesive.

The current work describes a Smalltalk implementation of the basic system with
the aim to learn and experiment with a pure object environment, such as Pharo
and Squeak. The work \citep{evan:smallkanren} is another Smalltalk
implementation of $\mu$Kanren that currently supports disequality constraints
and nominal logic; it is a more complete and complex version that we point it
out for users requiring those features -- hoping to be able to provide those
features by ourselves.

Starting small, the next example shows three different
approaches to define the addition of three \verb|SmallInteger| objects toward 
the introduction of the relational paradigm.
\begin{example}
In math a relation $P$ is usually characterized by a statement such as
\begin{displaymath}
\forall a,b,c.\,P(a,b,c) \leftrightarrow a + b = c \quad\text{from which}\quad
P(1,2,3) \quad\text{can be proved.}
\end{displaymath}
The same check can be expressed in Smalltalk using either the
\textit{imperative style}
\begin{minted}[fontsize=\footnotesize]{smalltalk}
a := 1.  b := 2.  c := a + b.  Object assert: [ c = 3 ]
\end{minted}
or the \textit{functional style}
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Object assert: [ ([ :a :b | a + b ] value: 1 value: 2) = 3 ]
\end{minted}
or, finally, the \textit{declarative style}
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Object assert: [ [ :a :b :c | a + b = c ] value: 1 value: 2 value: 3 ]
\end{minted}
which closely resembles the mathematical statement.
\end{example}

### 2019-03-14 booklet/article.tex
Describe the walking process and update the Chain hierarchy.

@@
 
-\subsection{Goals and substitutions}
+\subsection{\texttt{Goal}s and \texttt{State}s}
 
@@
 
-\subsection{Chains}
+\subsection{\texttt{Chain}s and \texttt{Cons}es}
 
@@
 \begin{minted}[fontsize=\footnotesize]{smalltalk}
+ChainTest>>#collatz: o
+    ^ o
+        cons: [o even
+                ifTrue: [self collatz: o / 2]
+                ifFalse: [self collatz: 3 * o + 1]]
+
 ChainTest>>#fibs
@@
 ChainTest>>#fib: m fib: n
-    ^ [self fib: n fib: m + n] links: m
+    ^ m cons: [self fib: n fib: m + n]
 
 ChainTest>>#gcd: m gcd: n
-    ^ [n isZero 
+    ^ m cons: [n isZero
                 ifTrue: [Chain with: m]
                 ifFalse: [self gcd: n gcd: m \\ n]]
-        links: m
 
@@
 ChainTest>>#ints: i
-    ^ [self ints: i + 1] links: i
+    ^ i cons: [self ints: i + 1]
 
-ChainTest>>#upto2
-    ^ [[[Chain bottom] links: 2] links: 1] links: 0
+ChainTest>>#zeroOneTwo
+    ^ 0 cons: (1 cons: (2 cons: Chain bottom))
 \end{minted}
-where blocks define the message \Verb|#links:| as 
+where any \Verb|Object| object respond to the message
 \begin{minted}[fontsize=\footnotesize]{smalltalk}
-BlockClosure>>#links: anObj
-    ^ Chain item: anObj linker: self
+Object>>#cons: anObj
+    ^ anObj consedObject: self
 \end{minted}
-exposing the abstract class \Verb|Chain| and its class-side messages
+which dispatches over
+\begin{minted}[fontsize=\footnotesize]{smalltalk}
+Object>>#consedObject: car
+    ^ Cons car: car asCons cdr: self asCons
+
+BlockClosure>>#consedObject: car
+    ^ Chain item: car linker: self
+
+Chain>>#consedObject: car
+    ^ car cons: [self]
+\end{minted}
+where (i)~\Verb|Cons| objects mimic \textit{Lisp's cons cells}
+\begin{minted}[fontsize=\footnotesize]{smalltalk}
+Cons class>>#car: anObj cdr: anotherObj 
+    ^ self new 
+        car: anObj;
+        cdr: anotherObj;
+        yourself
+
+Object>>#asCons
+    ^ Cons fromObject: self
+
+Cons class>>#fromObject: anObj 
+    ^ anObj
+\end{minted}
+and (ii)~the abstract class \Verb|Chain| provides the following class-side messages
 \begin{minted}[fontsize=\footnotesize]{smalltalk}
@@
 Chain class>>#with: anObj
-    ^ [Chain bottom] links: anObj
+    ^ anObj cons: Chain bottom
 
 Chain class>>#repeat: anObj
-    ^ [self repeat: anObj] links: anObj
+    ^ anObj cons: [self repeat: anObj]
 \end{minted}
-together with its two subclasses \Verb|Knot| and \Verb|Bottom|; a comprehensive
+together with its two subclasses \Verb|Knot| and \Verb|Bottom|. A comprehensive
 test case follows
@@
     self
-        assert: (self upto2 next: 10) contents
+        assert: (self zeroOneTwo next: 10) contents
         equals: (0 to: 2).
     self
-        assert: (self upto2 next: 1) contents
+        assert: (self zeroOneTwo next: 1) contents
         equals: (0 to: 0).
@@
 \end{minted}
+
 On one hand, when a \Verb|Chain| object (i)~receives the selector
@@
 \end{minted}
+
 On the other hand, when a \Verb|Chain| object (ii)~receives the selector
@@
 \end{minted}
-Observe that the \Verb|Chain| hierarchy support part of the protocol provided
-by\Verb|ReadStream| objects because of the messages \Verb|#next|, \Verb|#next:|
+Observe that the \Verb|Chain| hierarchy support the protocol provided
+by \Verb|ReadStream| objects because of the messages \Verb|#next|, \Verb|#next:|
 and \Verb|#contents|.
+
+\subsection{\texttt{Runner}s, \texttt{Reifier}s and \texttt{Var}s}
+
+Recall that a goal $g$ responds with a \Verb|Chain| that contains either zero,
+one, or more substitutions when it receives the message \Verb|#onState:|; in order
+to manipulate those substitutions, $g$ provides the message
+\begin{minted}[fontsize=\footnotesize]{smalltalk}
+Goal>>#solutions
+    ^ Runner onGoal: self
+
+Runner class>>#onGoal: aGoal 
+    ^ self new 
+        goal: aGoal;
+        yourself
+\end{minted}
+which builds a \Verb|Runner| object able to process the sequence of substitution, both (i)~all of them with the message
+\begin{minted}[fontsize=\footnotesize]{smalltalk}
+Runner>>#all
+    | chain |
+    chain := goal onState: self emptyState.
+    ^ self reify: chain contents
+\end{minted}
+and (ii)~at most $n$ of them with the message
+\begin{minted}[fontsize=\footnotesize]{smalltalk}
+Runner>>#atMost: anInteger 
+    | chain |
+    chain := goal onState: self emptyState.
+    ^ self reify: (chain next: anInteger) contents
+\end{minted}
+using the helper message 
+\begin{minted}[fontsize=\footnotesize]{smalltalk}
+Runner>>#emptyState
+    ^ Dictionary new asState
+\end{minted}
+
+We are now ready to describe the reification process which starts in
+\begin{minted}[fontsize=\footnotesize]{smalltalk}
+Runner>>#reify: states 
+    | vars |
+    vars := goal vars ifEmpty: [{Var tautology}].
+    ^ states
+        collect: [:s | "`s` stands for `state`"
+                    | rs | "`rs` stands for `reified state`"
+                    rs := s reify: vars.
+                    rs size = 1 ifTrue: [rs first] ifFalse: [rs]]
+\end{minted}
+where the accessor \Verb|#vars| is defined for any \Verb|Goal| objects and overridden by \Verb|Fresh| goals 
+\begin{minted}[fontsize=\footnotesize]{smalltalk}
+Goal>>#vars
+    ^ {}
+
+Fresh>>#vars
+    ^ vars
+\end{minted}
+respectively; in parallel, every \Verb|State| object responds to the selector \Verb|#reify:| with
+\begin{minted}[fontsize=\footnotesize]{smalltalk}
+State>>#reify: vars 
+    | walker reifier |
+    walker := Walker on: self.
+    reifier := Reifier on: Dictionary new asState.
+    ^ vars
+        collect: [:v | 
+                    | w s |
+                    w := walker walkStar: v.
+                    s := reifier reify: w.
+                    (Walker on: s) walkStar: w]
+\end{minted}
+where a \Verb|Walker| object is built by 
+\begin{minted}[fontsize=\footnotesize]{smalltalk}
+Walker class>>#on: aState
+    ^ self new 
+        state: aState;
+        yourself
+\end{minted}
+and it responds to the selector \Verb|#walkStar| with
+\begin{minted}[fontsize=\footnotesize]{smalltalk}
+Walker>>#walkStar: aVar 
+    ^ (state walk: aVar)
+        walkStarUsingWalker: self
+\end{minted}
+which relies on 
+\begin{minted}[fontsize=\footnotesize]{smalltalk}
+State>>#walk: anObj 
+    | k |
+    k := anObj.
+    [k := substitution
+            at: k
+            ifAbsent: [^ k]] repeat
+\end{minted}
+and different objects responds to \Verb|#walkStarUsingWalker:| as follows
+\begin{minted}[fontsize=\footnotesize]{smalltalk}
+Object>>#walkStarUsingWalker: aWalker
+    ^ aWalker walkStarObject: self
+
+Array>>#walkStarUsingWalker: aWalker
+    ^ aWalker walkStarArray: self
+
+LinkedList>>#walkStarUsingWalker: aWalker
+    ^ aWalker walkStarLinkedList: self
+
+Cons>>#walkStarUsingWalker: aWalker
+    ^ aWalker walkStarCons: self 
+\end{minted}
+dispatching over \Verb|Walker| objects
+\begin{minted}[fontsize=\footnotesize]{smalltalk}
+Walker>>#walkStarObject: anObj 
+    ^ state walk: anObj
+
+Walker>>#walkStarArray: aCollection 
+    ^ aCollection
+        collect: [:each | self walkStar: each]
+
+Walker>>#walkStarLinkedList: aCollection 
+    ^ aCollection
+        collect: [:each | self walkStar: each]
+
+Walker>>#walkStarCons: aCons 
+    | car cdr |
+    car := self walkStar: aCons car.
+    cdr := self walkStar: aCons cdr.
+    ^ car cons: cdr 
+\end{minted}
+respectively.
 

### 2019-03-16 booklet/article.tex
Add a preliminary description of the reification process.

@@
 goal\footnote{put here the reference to its subsection}. An empty substitution
-is created by \stcode{Dictionary new asState} where
+is created by
+\begin{minted}[fontsize=\footnotesize]{smalltalk}
+Dictionary new asState
+\end{minted}
+where
 \begin{minted}[fontsize=\footnotesize]{smalltalk}
@@
 \end{minted}
-mantaining a protocol coherent to the one of \Verb|Dictionary| objects for
-extension.
+mantaining a coherent protocol with respect to the one of \Verb|Dictionary|
+objects for extension; moreover, \Verb|UnificationError| objects will be
+explained in a later section when will talk about the unification process.
 
@@
 \end{minted}
-together with its two subclasses \Verb|Knot| and \Verb|Bottom|. A comprehensive
+having two subclasses \Verb|Knot| and \Verb|Bottom|. A comprehensive
 test case follows
@@
 On one hand, when a \Verb|Chain| object (i)~receives the selector
-\Verb|#next:| that carries a \Verb|SmallInteger| object $i$ it keeps at most
-$i$ objects and then truncate itself, precisely
+\Verb|#next:| that carries a \Verb|SmallInteger| object $n$ then it keeps at most
+$n$ objects and then truncate itself, precisely
 \begin{minted}[fontsize=\footnotesize]{smalltalk}
-Bottom>>#next: anInteger
+Bottom>>#next: n
     ^ self
@@
 Knot>>#next: n
-    ^ n isZero
-        ifTrue: [Chain bottom]
-        ifFalse: [[self next value next: n - 1] links: item]
+    ^ n = 1
+        ifTrue: [Chain with: item]
+        ifFalse: [item cons: [self next next: n - 1]]
 \end{minted}
@@
 
-Recall that a goal $g$ responds with a \Verb|Chain| that contains either zero,
-one, or more substitutions when it receives the message \Verb|#onState:|; in order
-to manipulate those substitutions, $g$ provides the message
+Recall that a \Verb|Goal| object responds with a \Verb|Chain| that contains either zero,
+one, or more substitutions when it receives the message \Verb|#onState:| and it
+delegates the task to process those substitutions to \Verb|Runner| objects,
 \begin{minted}[fontsize=\footnotesize]{smalltalk}
@@
 \end{minted}
-which builds a \Verb|Runner| object able to process the sequence of substitution, both (i)~all of them with the message
+which are able to process both (i)~all of them with the message
 \begin{minted}[fontsize=\footnotesize]{smalltalk}
 Runner>>#all
-    | chain |
-    chain := goal onState: self emptyState.
-    ^ self reify: chain contents
+    ^ self run contents
 \end{minted}
@@
 Runner>>#atMost: anInteger
-    | chain |
-    chain := goal onState: self emptyState.
-    ^ self reify: (chain next: anInteger) contents
+    ^ (self run next: anInteger) contents
 \end{minted}
-using the helper message 
+relying on the unary message
 \begin{minted}[fontsize=\footnotesize]{smalltalk}
-Runner>>#emptyState
-    ^ Dictionary new asState
+Runner>>#run
+    | states vars selector |
+    states := goal onState: Dictionary new asState.
+    vars := goal vars ifEmpty: [LinkedList with: Var tautology].
+    selector := vars size = 1 ifTrue: #first ifFalse: #yourself.
+    ^ states
+        collect: [:s | selector
+                        value: (s reifier reifyVars: vars)]
 \end{minted}
+having the responsibilities
+\begin{itemize}
 
-We are now ready to describe the reification process which starts in
-\begin{minted}[fontsize=\footnotesize]{smalltalk}
-Runner>>#reify: states 
-    | vars |
-    vars := goal vars ifEmpty: [{Var tautology}].
-    ^ states
-        collect: [:s | "`s` stands for `state`"
-                    | rs | "`rs` stands for `reified state`"
-                    rs := s reify: vars.
-                    rs size = 1 ifTrue: [rs first] ifFalse: [rs]]
-\end{minted}
-where the accessor \Verb|#vars| is defined for any \Verb|Goal| objects and overridden by \Verb|Fresh| goals 
+\item to send the message \Verb|#onState:| to \Verb|Goal| objects to start the
+resolution process from an empty state, this can be considered the entry point of
+the ``computation'' that produces a \Verb|Chain| of substitutions;
+
+\item to take into account the variables with respect to results should be
+provided; as a particular case, if the root goal doesn't introduce any logic
+variable, by means of the auxiliary messages
 \begin{minted}[fontsize=\footnotesize]{smalltalk}
@@
 \end{minted}
-respectively; in parallel, every \Verb|State| object responds to the selector \Verb|#reify:| with
+then it is equivalent to prove a \textit{tautology}, represented by
 \begin{minted}[fontsize=\footnotesize]{smalltalk}
-State>>#reify: vars 
+Var class>>#tautology
+    ^ #tautology
+\end{minted}
+which require to decide that a relation is always true regardless values taken
+by logic variables;
+
+\item to simplify the results presentation when the root \Verb|Goal| object
+introduces \textit{exactly one} logic variable: in this case a
+\Verb|LinkedList| of values for that variable will be produced; otherwise, a
+\Verb|LinkedList| of \Verb|LinkedList| of values will be produces each one for
+each tuple of logic variables introduced by the root \Verb|Goal| object -- this
+point will be clear when we'll explain \Verb|Fresh| goals that are the only
+ones able to introduce logic variables;
+
+\item to map each substitution (always translate to \Verb|State| object) that
+satisfies the root \Verb|Goal| object do \textit{reify} the required variables,
+this is accomplished by the messages
+\begin{minted}[fontsize=\footnotesize]{smalltalk}
+s reifier reifyVars: vars
+\end{minted}
+within the block argument of the returning \Verb|#collect:| message send.
+\end{itemize}
+The last point shows that \Verb|Chain| objects are \textit{functors} according
+to the Haskell terminology because they answer to the message \Verb|#collect:|,
+which is a synonism for the function \Verb|fmap :: (a -> b) -> f a -> f b|, as
+\begin{minted}[fontsize=\footnotesize]{smalltalk}
+Chain>>#collect: aBlockClosure
+    ^ self
+
+Knot>>#collect: aBlockClosure
+    | v |
+    v := aBlockClosure value: item.
+    ^ v cons: [self next collect: aBlockClosure]
+\end{minted}
+respectively.  We are now ready to describe the reification process which
+requires
+\begin{minted}[fontsize=\footnotesize]{smalltalk}
+State>>#reifier
+    ^ Reifier on: self
+
+Reifier>>#reifyVars: vars
     | walker reifier |
-    walker := Walker on: self.
-    reifier := Reifier on: Dictionary new asState.
+    walker := state walker.
+    reifier := Dictionary new asState reifier.
     ^ vars
@@
                     s := reifier reify: w.
-                    (Walker on: s) walkStar: w]
+                    s walker walkStar: w]
 \end{minted}
+which uses the message
+\begin{minted}[fontsize=\footnotesize]{smalltalk}
+Reifier>>#reify: anObj 
+    ^ (state walk: anObj) 
+        reifyUsing: self
+\end{minted}
+to dispatch the message \Verb|#reifyUsing:| over
+\begin{minted}[fontsize=\footnotesize]{smalltalk}
+Object>>#reifyUsing: aReifier
+    ^ aReifier reifyObject: self
+
+Var>>#reifyUsing: aReifier 
+    ^ aReifier reifyVar: self
+
+ReifiedVar>>#reifyUsing: aReifier 
+    ^ aReifier reifyReifiedVar: self
+
+Array>>#reifyUsing: aReifier
+    ^ aReifier reifyArray: self
+
+LinkedList>>#reifyUsing: aReifier 
+    ^ aReifier reifyLinkedList: self
+
+Cons>>#reifyUsing: aReifier 
+    ^ aReifier reifyCons: self 
+\end{minted}
+and to collect replies as
+\begin{minted}[fontsize=\footnotesize]{smalltalk}
+Reifier>>#reifyObject: anObj 
+    ^ state
+
+Reifier>>#reifyVar: aVar 
+    ^ state reifyVar: aVar
+
+Reifier>>#reifyReifiedVar: aVar 
+    ^ state
+
+Reifier>>#reifyArray: aCollection 
+    ^ self reifyCollection: aCollection
+
+Reifier>>#reifyArray: aCollection 
+    ^ self reifyCollection: aCollection
+
+Reifier>>#reifyArray: aCollection 
+    ^ self reifyCollection: aCollection
+
+Reifier>>#reifyCollection: aCollection 
+    ^ aCollection inject: state into: [:s :c | s reifier reify: c]
+\end{minted}
+requiring the auxiliary message
+\begin{minted}[fontsize=\footnotesize]{smalltalk}
+State>>#reifyVar: aVar 
+    substitution
+        at: aVar
+        ifAbsentPut: [ReifiedVar new 
+                        id: substitution size; 
+                        yourself]
+\end{minted}
+and
+\begin{minted}[fontsize=\footnotesize]{smalltalk}
+State>>#walk: anObj 
+    | k |
+    k := anObj.
+    [k := substitution at: k ifAbsent: [^ k]] repeat
+\end{minted}
+
+\hrulefill
+
 where a \Verb|Walker| object is built by
@@
 \end{minted}
-and different objects responds to \Verb|#walkStarUsingWalker:| as follows
+and objects of different classes responds to \Verb|#walkStarUsingWalker:| as follows
 \begin{minted}[fontsize=\footnotesize]{smalltalk}

### 2019-03-17 booklet/article.tex
a first pass

@@
-\documentclass[a4paper,12pt]{article}
+\documentclass[a4paper,11pt]{article}
 
@@
 
+% ********************************************************************* 
+% Revisions and comments Macros
+% ********************************************************************* 
+\usepackage{xcolor}
+\usepackage[normalem]{ulem} 			% \sout macro
+\newboolean{showcomments}
+% \setboolean{showcomments}{false}
+\setboolean{showcomments}{true}
+
+\ifthenelse{\boolean{showcomments}}
+{
+	\newcommand{\nb}[3]{
+		{\colorbox{#2}{\bfseries\sffamily\scriptsize\textcolor{white}{#1}}}
+		{\textcolor{#2}{\sf\small$\blacktriangleright$\textit{#3}$\blacktriangleleft$}}}
+	\newcommand{\bnote}[2]{\fbox{\color{blue}\bfseries\sffamily\scriptsize#1}
+    	{\color{blue}\sf\small$\blacktriangleright$\textit{#2}$\blacktriangleleft$}}
+	\newcommand{\old}[1]{{\color{gray}\sout{#1}}} % old to be removed
+	\newcommand{\del}[1]{\old{#1}} % please remove
+	\newcommand{\ins}[1]{{\textcolor{blue}{\uline{#1}}}} % please insert	
+	\newcommand{\ugh}[1]{{\textcolor{red}{\uwave{#1}}}} % please rephrase	
+	\newcommand{\chg}[2]{{\textcolor{red}{\sout{#1}}}{\ra}\textcolor{blue}{\uline{#2}}} % please change
+	\newcommand{\here}{\bnote{***}{CONTINUE HERE}} 
+	\newcommand{\fix}[1]{\bnote{FIX}{#1}}
+}{
+	\newcommand{\bnote}[2]{}
+	\newcommand{\nb}[3]{}
+	\newcommand{\old}[1]{}
+	\newcommand{\del}[1]{}
+	\newcommand{\ins}[1]{}
+	\newcommand{\ugh}[1]{}
+	\newcommand{\chg}[2]{}
+	\newcommand{\here}{}
+	\newcommand{\fix}[1]{}
+} 
+%---% add your own macros 
+\newcommand{\sd}[1]{\nb{Stef}{orange}{#1}}
+
+\newcommand{\ct}[1]{{\textsf{#1}}\xspace}
+
 \author{Massimo Nocentini}
@@
 
-\subsection{\texttt{Goal}s and \texttt{State}s}
+\sd{As a style point of view I would not put code inside sentence. We should be able to read the text without the code and the code without the text. It makes paragraphs more self contained.}
 
-A \textit{goal} is any object that responds to the selector \Verb|#onState:| so
+\sd{Why don't you start with Cons. Then once people understand Cons and infinite stream
+you can introduce Goal and state and always example first.}
+
+\sd{Having an diagram showing how an object will be composed will also help}
+
+\subsection{\ct{Goal}s and \ct{State}s}
+
+A \textit{goal} is any object that responds to the message \ct{\#onState:} so
 that it receives a substitution and produces an ordered sequence of zero or
-more substitutions. In turn, a substitution is an unordered collection of
+more substitutions.
+
+\sd{Show a test. I do not get what is a state? A variable? Why this is call onState:}
+\sd{Why this is not \ct{goal withSubstitution: aSubstitution}}
+
+ In turn, a substitution is an unordered collection of
 values which are indexed by \textit{logic variables}, therefore a
-\Verb|Dictionary| object fits perfectly; we wrap it in a \Verb|State| object
-that also keeps record of an incremental numbering \Verb|birthdate| whose
-meaning will be clarified later when we'll describe the \Verb|Fresh|
+\ct{Dictionary} object fits perfectly \sd{show an example now}; we wrap it \sd{who?} in a \ct{State} object
+that also keeps record of an incremental numbering  \ct{birthdate} whose
+meaning will be clarified later when we'll describe the  \ct{Fresh}
 goal\footnote{put here the reference to its subsection}. An empty substitution
@@
 \begin{minted}[fontsize=\footnotesize]{smalltalk}
-Dictionary>>#asState
+Dictionary >> asState
     ^ State new
@@
 \begin{minted}[fontsize=\footnotesize]{smalltalk}
-State>>#at: aVar put: aValue
+State >> at: aVar put: aValue
     | s |
     s := substitution copy.
-    s "check that if a key already exists at aVar, then the values must be the same"
+    "check that if a key already exists at aVar, then the values must be the same"
+    s 
       at: aVar
@@
 \end{minted}
-mantaining a coherent protocol with respect to the one of \Verb|Dictionary|
-objects for extension; moreover, \Verb|UnificationError| objects will be
+maintaining a coherent protocol with respect to the one of \Verb|Dictionary|
+objects for extension; moreover, \ct{UnificationError} objects will be
 explained in a later section when will talk about the unification process.
 
-\subsection{\texttt{Chain}s and \texttt{Cons}es}
+\subsection{\ct{Chain}s and \ct{Cons}es}
+\sd{As a general comments. You should say to the reader. We will do that. And do piece by piece
+Now I discover things just stumbling on them.}
+
+\sd{Example: In this section we will build the infrastructure to represent potentially infinite lists. 
+For this will start building a chain. A chain is composed of consed: objects following Lisp tradition. 
+We will define first the class Chain as an abstract class with two subclass \ct{Knot} and \ct{Bottom}.
+\ct{Bottom} will represent the empty cell of lisp on top of which any list is build. \ct{Knot} represents 
+a ...
+}
 
 Because the sequence of substitutions may be infinite, we
-represent it not as a \Verb|LinkedList| object but as a \Verb|Chain| object;
-as usual, we find \Verb|BlockClosure| objects good candidates to represent (a possibly
-infinite) computation. Some examples are in order,
+represent it not as a \ct{LinkedList} object but as a \ct{Chain} object;
+as usual, we find \ct{BlockClosure} objects good candidates to represent (a possibly
+infinite) computation. Some examples are in order, \sd{ChainTest -> ChainExamples}
 \begin{minted}[fontsize=\footnotesize]{smalltalk}
-ChainTest>>#collatz: o
+ChainTest >> collatz: o
     ^ o
@@
 
-ChainTest>>#fibs
+ChainTest >> fibs
     ^ self fib: 0 fib: 1
 
-ChainTest>>#fib: m fib: n
+ChainTest >> fib: m fib: n
     ^ m cons: [self fib: n fib: m + n]
 
-ChainTest>>#gcd: m gcd: n
+ChainTest >> gcd: m gcd: n
     ^ m cons: [n isZero
@@
 
-ChainTest>>#nats
+ChainTest >> nats
     ^ self ints: 0
 
-ChainTest>>#ints: i
+ChainTest >> ints: i
     ^ i cons: [self ints: i + 1]
 
-ChainTest>>#zeroOneTwo
+ChainTest >> zeroOneTwo
     ^ 0 cons: (1 cons: (2 cons: Chain bottom))
 \end{minted}
-where any \Verb|Object| object respond to the message
+where any \ct{Object} object respond to the message \ct{cons:} as follows: 
+\sd{what is the argument?}
+\sd{Separate Cons and Chain explanation}
+\sd{why not just started with Cons. Ok we have an object}
 \begin{minted}[fontsize=\footnotesize]{smalltalk}
-Object>>#cons: anObj
+Object >> cons: anObj
     ^ anObj consedObject: self
@@
 \begin{minted}[fontsize=\footnotesize]{smalltalk}
-Object>>#consedObject: car
+Object >> consedObject: car
     ^ Cons car: car asCons cdr: self asCons
 
-BlockClosure>>#consedObject: car
+BlockClosure >> consedObject: car
     ^ Chain item: car linker: self
 
-Chain>>#consedObject: car
+Chain >> consedObject: car
     ^ car cons: [self]
 \end{minted}
-where (i)~\Verb|Cons| objects mimic \textit{Lisp's cons cells}
+where (i)~\ct{Cons} objects mimic \textit{Lisp's cons cells}
 \begin{minted}[fontsize=\footnotesize]{smalltalk}
-Cons class>>#car: anObj cdr: anotherObj
+Cons class >>  car: anObj cdr: anotherObj
     ^ self new
@@
 
-Object>>#asCons
+Object >> asCons
     ^ Cons fromObject: self
 
-Cons class>>#fromObject: anObj
+Cons class >> fromObject: anObj
     ^ anObj
 \end{minted}
-and (ii)~the abstract class \Verb|Chain| provides the following class-side messages
+
+\s{I do not get fromObject: implementation. because you do not wrap the object.
+So 1 asCons will return 1}
+and (ii)~the abstract class \ct{Chain} provides the following class-side messages
 \begin{minted}[fontsize=\footnotesize]{smalltalk}
-Chain class>>#bottom
+Chain class >> bottom
     ^ Bottom new
 
-Chain class>>#item: anObj linker: aBlockClosure
+Chain class >> item: anObj linker: aBlockClosure
     ^ Knot new
@@
 
-Chain class>>#with: anObj
+Chain class >> with: anObj
     ^ anObj cons: Chain bottom
 
-Chain class>>#repeat: anObj
+Chain class >> repeat: anObj
     ^ anObj cons: [self repeat: anObj]
 \end{minted}
-having two subclasses \Verb|Knot| and \Verb|Bottom|. A comprehensive
-test case follows
+having two subclasses \ct{Knot} and \ct{Bottom}. A comprehensive
+test case follows \sd{split it and explain.}
 \begin{minted}[fontsize=\footnotesize]{smalltalk}
-ChainTest>>#testNumbers
+ChainTest >> testNumbers
     self
@@
 
-On one hand, when a \Verb|Chain| object (i)~receives the selector
-\Verb|#next:| that carries a \Verb|SmallInteger| object $n$ then it keeps at most
+On one hand, when a \ct{Chain} object (i)~receives the message
+\ct{next:} that carries a \ct{SmallInteger} object $n$ then it keeps at most
 $n$ objects and then truncate itself, precisely
 \begin{minted}[fontsize=\footnotesize]{smalltalk}
-Bottom>>#next: n
+Bottom >> next: n
     ^ self
 
-Knot>>#next: n
+Knot >> next: n
     ^ n = 1
@@
 
-On the other hand, when a \Verb|Chain| object (ii)~receives the selector
-\Verb|#contents| it builds a \Verb|LinkedList| object \textit{eagerly},
+On the other hand, when a \ct{Chain} object (ii)~receives the message
+\ct{#contents} it builds a \ct{LinkedList} object \textit{eagerly},
 \begin{minted}[fontsize=\footnotesize]{smalltalk}
-Bottom>>#contents
+Bottom >> contents
     ^ LinkedList new
 
-Knot>>#contents
+Knot >> contents
     ^ self next contents
@@
 \end{minted}
-Observe that the \Verb|Chain| hierarchy support the protocol provided
-by \Verb|ReadStream| objects because of the messages \Verb|#next|, \Verb|#next:|
-and \Verb|#contents|.
+Observe that the \ct{|Chain} hierarchy supports the protocol provided
+by \ct{ReadStream} objects because of the messages \ct{#next}, \ct{#next:}
+and \ct{#contents}.
 
@@
 
-Recall that a \Verb|Goal| object responds with a \Verb|Chain| that contains either zero,
-one, or more substitutions when it receives the message \Verb|#onState:| and it
-delegates the task to process those substitutions to \Verb|Runner| objects,
+Recall \sd{it was not said that way so I do not recall} that a \ct{Goal} object responds with a \ct{Chain} that contains either zero,
+one, or more substitutions when it receives the message \ct{#onState:} and it
+delegates the task to process those substitutions to \ct{Runner} objects,
+
 \begin{minted}[fontsize=\footnotesize]{smalltalk}
-Goal>>#solutions
+Goal >> solutions
     ^ Runner onGoal: self
 
-Runner class>>#onGoal: aGoal
+Runner class >> onGoal: aGoal
     ^ self new
@@
 \begin{minted}[fontsize=\footnotesize]{smalltalk}
-Runner>>#all
+Runner >> all
     ^ self run contents
 \end{minted}
-and (ii)~at most $n$ of them with the message
+and (ii)~at most $n$ of them with the message \ct{atMost:} below: 
 \begin{minted}[fontsize=\footnotesize]{smalltalk}
-Runner>>#atMost: anInteger
+Runner >> atMost: anInteger
     ^ (self run next: anInteger) contents
 \end{minted}
-relying on the unary message
+relying on the unary message \ct{run}
+
 \begin{minted}[fontsize=\footnotesize]{smalltalk}

### 2019-03-18 booklet/article.tex
Update and add a preliminary version of the walking process.

--- 2019-03-16 booklet/article.tex
@@
 
-Reifier>>#reifyArray: aCollection 
+Reifier>>#reifyLinkedList: aCollection 
     ^ self reifyCollection: aCollection
 
-Reifier>>#reifyArray: aCollection 
+Reifier>>#reifyCons: aCollection 
     ^ self reifyCollection: aCollection
@@
 \end{minted}
+Observe that \Verb|Chain| objects are also \textit{foldable} according to the Haskell \Verb|Foldable| typeclass because 
+they respond to the message
+\begin{minted}[fontsize=\footnotesize]{smalltalk}
+Cons>>#inject: init into: aBlock 
+    | accumulated |
+    accumulated := aBlock value: init value: car.
+    ^ aBlock value: accumulated value: cdr
+\end{minted}
 
-\hrulefill
+The reification process requires \Verb|Walker| objects which are built by
+\Verb|State| objects using the template method
+\begin{minted}[fontsize=\footnotesize]{smalltalk}
+State>>#walker
+    ^ Walker on: self
 
-where a \Verb|Walker| object is built by
-\begin{minted}[fontsize=\footnotesize]{smalltalk}
 Walker class>>#on: aState
@@
 \end{minted}
-and it responds to the selector \Verb|#walkStar| with
+\Verb|Walker| responds to the selector \Verb|#walkStar| with
 \begin{minted}[fontsize=\footnotesize]{smalltalk}
@@
         walkStarUsingWalker: self
-\end{minted}
-which relies on
-\begin{minted}[fontsize=\footnotesize]{smalltalk}
-State>>#walk: anObj
-    | k |
-    k := anObj.
-    [k := substitution
-            at: k
-            ifAbsent: [^ k]] repeat
 \end{minted}

### 2019-03-18 booklet/article.tex
Little fixes to compile with make.

--- 2019-03-17 booklet/article.tex
@@
 \usepackage{xcolor}
+\usepackage{xspace}
 \usepackage[normalem]{ulem} 			% \sout macro
@@
 \newcommand{\sd}[1]{\nb{Stef}{orange}{#1}}
+\newcommand{\mn}[1]{\nb{Massimo}{blue}{#1}}
 
@@
 
-\s{I do not get fromObject: implementation. because you do not wrap the object.
+\sd{I do not get fromObject: implementation. because you do not wrap the object.
 So 1 asCons will return 1}
@@
 On the other hand, when a \ct{Chain} object (ii)~receives the message
-\ct{#contents} it builds a \ct{LinkedList} object \textit{eagerly},
+\ct{\#contents} it builds a \ct{LinkedList} object \textit{eagerly},
 \begin{minted}[fontsize=\footnotesize]{smalltalk}
@@
 \end{minted}
-Observe that the \ct{|Chain} hierarchy supports the protocol provided
-by \ct{ReadStream} objects because of the messages \ct{#next}, \ct{#next:}
-and \ct{#contents}.
+Observe that the \ct{Chain} hierarchy supports the protocol provided
+by \ct{ReadStream} objects because of the messages \ct{\#next}, \ct{\#next:}
+and \ct{\#contents}.
 
@@
 Recall \sd{it was not said that way so I do not recall} that a \ct{Goal} object responds with a \ct{Chain} that contains either zero,
-one, or more substitutions when it receives the message \ct{#onState:} and it
+one, or more substitutions when it receives the message \ct{\#onState:} and it
 delegates the task to process those substitutions to \ct{Runner} objects,

### 2019-03-18 booklet/article.tex
Restore some paragraphs deleted by the merge, my fault.

@@
 \end{minted}
+Observe that \ct{Chain} objects are also \textit{foldable} according to the
+Haskell \ct{Foldable} typeclass because they respond to the message
+\begin{minted}[fontsize=\footnotesize]{smalltalk}
+Cons>>#inject: init into: aBlock 
+    | accumulated |
+    accumulated := aBlock value: init value: car.
+    ^ aBlock value: accumulated value: cdr
+\end{minted}
 
-\hrulefill
-
-where a \Verb|Walker| object is built by
+The reification process requires \ct{Walker} objects which are built by
+\ct{State} objects using the template method
 \begin{minted}[fontsize=\footnotesize]{smalltalk}
-Walker class>>#on: aState
-    ^ self new
-        state: aState;
-        yourself
+State>>#walker
+    ^ Walker on: self
 \end{minted}

### 2019-03-19 booklet/article.tex
Update and apply suggestions by Stef Ducasse.

@@
 
-\sd{As a style point of view I would not put code inside sentence. We should be able to read the text without the code and the code without the text. It makes paragraphs more self contained.}
+\sd{As a style point of view I would not put code inside sentence. We should be
+able to read the text without the code and the code without the text. It makes
+paragraphs more self contained.}
 
 \sd{Why don't you start with Cons. Then once people understand Cons and infinite stream
-you can introduce Goal and state and always example first.}
+you can introduce Goal and state and always example first.
+\mn{I do prefer to explain both of them when the need arises to not postpone the main definition of Goal.}}
 
@@
 
-\subsection{\ct{Goal}s and \ct{State}s}
+In this section we describe the fundamentals of $\mu$Kanren and the protocols
+that objects provide to compose the architecture on which Section
+\ref{sec:bootstrapping} will define core logic operators.
 
-A \textit{goal} is any object that responds to the message \ct{\#onState:} so
-that it receives a substitution and produces an ordered sequence of zero or
-more substitutions.
+We start describing objects that (i)~encodes relational statements,
+(ii)~logical variables and (iii)~associations of values to variables such that
+a given predicate can be satisfied.
 
-\sd{Show a test. I do not get what is a state? A variable? Why this is call onState:}
-\sd{Why this is not \ct{goal withSubstitution: aSubstitution}}
+\subsection{Goals, variables and states}
+
+A \textit{goal} $g$ denotes a logical relation; precisely, $g$ is any object
+that responds to the message \ct{\#onState:} so that it receives a substitution
+and produces an ordered sequence of zero or more substitutions.
 
@@
 values which are indexed by \textit{logic variables}, therefore a
-\ct{Dictionary} object fits perfectly \sd{show an example now}; we wrap it \sd{who?} in a \ct{State} object
-that also keeps record of an incremental numbering  \ct{birthdate} whose
+\ct{Dictionary} object fits perfectly. For the sake of clarity, the predicate
+\begin{displaymath}
+inc(X, succ(succ(zero))) \quad\text{where}\quad Y = succ(X) \rightarrow inc(X, Y),
+\end{displaymath}
+is satisfied by the substitution
+\begin{minted}[fontsize=\footnotesize]{smalltalk}
+|X|
+X := (Var id: 0).
+Dictionary new
+    at: X put: 'succ(zero)';
+    yourself
+\end{minted}
+
+Moreover, logic variables are created by means of the message
+\begin{minted}[fontsize=\footnotesize]{smalltalk}
+Var class >> id: anInteger
+    ^ self new
+        id: ('var_' , anInteger asString) asSymbol;
+        yourself
+\end{minted}
+
+We wrap a substitution in a \ct{State} object that keeps record of metadata
+about the computation, for instance a \ct{State} has an integer \ct{birthdate}
+to track the (discrete) time when a logic variable has been introduced -- its
 meaning will be clarified later when we'll describe the \ct{Fresh}
-goal\footnote{put here the reference to its subsection}. An empty substitution
-is created by
+goal\footnote{put here the reference to its subsection}.
+
+An initial \ct{State} object is created by
 \begin{minted}[fontsize=\footnotesize]{smalltalk}
@@
 maintaining a coherent protocol with respect to the one of \Verb|Dictionary|
-objects for extension; moreover, \ct{UnificationError} objects will be
+objects for extension; finally, \ct{UnificationError} objects will be
 explained in a later section when will talk about the unification process.
 
-\subsection{\ct{Chain}s and \ct{Cons}es}
-\sd{As a general comments. You should say to the reader. We will do that. And do piece by piece
-Now I discover things just stumbling on them.}
+\subsection{Cons cells}
 
-\sd{Example: In this section we will build the infrastructure to represent potentially infinite lists. 
-For this will start building a chain. A chain is composed of consed: objects following Lisp tradition. 
-We will define first the class Chain as an abstract class with two subclass \ct{Knot} and \ct{Bottom}.
-\ct{Bottom} will represent the empty cell of lisp on top of which any list is build. \ct{Knot} represents 
-a ...
-}
+Because the sequence of substitutions may be infinite, we represent it not as a
+\ct{LinkedList} object but as a \ct{Chain} object which has the capability to
+represent potentially infinite sequences.  For this we will start building
+\ct{Cons} cells; following the Lisp tradition, we reproduce the usual behavior
+as the next test method shows:
+\begin{minted}[fontsize=\footnotesize]{smalltalk}
+testCons
+    self
+        assert: '(0)'
+        equals: (0 cons: nil) asString.
+    self
+        assert: '(0 1 2 3 4 5 6 7 8 9)'
+        equals: (0 to: 9) asCons asString.
+    self
+        assert: '((1 2) 3 4 (5 (6 8)) 7)'
+        equals: {{1. 2}. 3. 4. {5. {6. 8}}. 7} asCons asString.
 
-Because the sequence of substitutions may be infinite, we
-represent it not as a \ct{LinkedList} object but as a \ct{Chain} object;
-as usual, we find \ct{BlockClosure} objects good candidates to represent (a possibly
-infinite) computation. Some examples are in order, \sd{ChainTest -> ChainExamples}
+    "dotted pairs notation"
+    self
+        assert: '(0 . 1)'
+        equals: (0 cons: 1) asString.
+    self
+        assert: '(0 1 2 3 4 5 6 7 8 9 . 10)'
+        equals: ((0 to: 9) asConsWithCdr: 10) asString.
+    self
+        assert: '(((3 . 4) . 5) 6 7)'
+        equals: (((3 cons: 4) cons: 5) cons: (6 cons: (7 cons: nil))) asString
+\end{minted}
+We split its description of
+\begin{itemize}
+\item the creation message \ct{\#cons:} is defined as
+\begin{minted}[fontsize=\footnotesize]{smalltalk}
+Object >> cons: anObj
+    ^ anObj consedObject: self
+\end{minted}
+which dispatches the message \ct{\#consedObject:} on its argument to build
+\ct{Cons} cells with
+\begin{minted}[fontsize=\footnotesize]{smalltalk}
+Object >> consedObject: car
+    ^ Cons car: car cdr: self
+
+Cons class >>  car: anObj cdr: anotherObj
+    ^ self new
+        car: anObj;
+        cdr: anotherObj;
+        yourself
+\end{minted}
+
+\item the translation message \ct{\#asCons} is defined and overridden as
+\begin{minted}[fontsize=\footnotesize]{smalltalk}
+Object >> asCons
+    ^ Cons fromObject: self
+
+ByteSymbol >> asCons
+    ^ Cons fromByteSymbol: self
+
+SequenceableCollection >> asCons
+    ^ Cons fromSequenceableCollection: self
+\end{minted}
+in parallel, the class object \ct{Cons} responds as follows
+\begin{minted}[fontsize=\footnotesize]{smalltalk}
+Cons >> fromObject: anObj
+    ^ anObj
+
+Cons >> fromByteSymbol: aSymbol
+    ^ aSymbol
+
+Cons >> fromSequenceableCollection: aCollection
+    ^ self fromSequenceableCollection: aCollection lastCdr: nil
+
+Cons >> fromSequenceableCollection: aCollection lastCdr: aCdr
+    ^ aCollection
+        foldr: [:each :acc | self car: each asCons cdr: acc asCons]
+        init: aCdr
+\end{minted}
+Finally, to build the more general notion of \textit{dotted pairs},
+the message \ct{\#asConsWithCdr:} is defined and overridden as
+\begin{minted}[fontsize=\footnotesize]{smalltalk}
+Object >> asConsWithCdr: aCdr
+    ^ Cons fromObject: self lastCdr: aCdr
+
+SequenceableCollection >> asConsWithCdr: aCdr
+    ^ Cons fromSequenceableCollection: self lastCdr: aCdr
+\end{minted}
+which require
+\begin{minted}[fontsize=\footnotesize]{smalltalk}
+Cons >> fromObject: anObj lastCdr: aCdr
+    ^ self car: anObj cdr: aCdr
+\end{minted}
+
+\end{itemize}
+
+\subsection{Chains}
+
+We find \ct{BlockClosure} objects good candidates to
+represent a suspended (and possibly infinite) computation. The following
+auxiliary methods for the test case \ct{ChainTest} show examples of how
+\ct{BlockClosure}s can be used as arguments for the message \ct{\#cons:},
 \begin{minted}[fontsize=\footnotesize]{smalltalk}
@@
 \end{minted}
-where any \ct{Object} object respond to the message \ct{cons:} as follows: 
-\sd{what is the argument?}
-\sd{Separate Cons and Chain explanation}
-\sd{why not just started with Cons. Ok we have an object}
+In turn, when a \ct{BlockClosure} receives the dispatched message
 \begin{minted}[fontsize=\footnotesize]{smalltalk}
-Object >> cons: anObj
-    ^ anObj consedObject: self
-\end{minted}
-which dispatches over
-\begin{minted}[fontsize=\footnotesize]{smalltalk}
-Object >> consedObject: car
-    ^ Cons car: car asCons cdr: self asCons
-
 BlockClosure >> consedObject: car
     ^ Chain item: car linker: self
-
-Chain >> consedObject: car
-    ^ car cons: [self]
 \end{minted}
-where (i)~\ct{Cons} objects mimic \textit{Lisp's cons cells}
-\begin{minted}[fontsize=\footnotesize]{smalltalk}
-Cons class >>  car: anObj cdr: anotherObj
-    ^ self new
-        car: anObj;
-        cdr: anotherObj;
-        yourself
-
-Object >> asCons
-    ^ Cons fromObject: self
-
-Cons class >> fromObject: anObj
-    ^ anObj
-\end{minted}
-
-\sd{I do not get fromObject: implementation. because you do not wrap the object.
-So 1 asCons will return 1}
-and (ii)~the abstract class \ct{Chain} provides the following class-side messages
+then it builds a \ct{Chain}; precisely, \ct{Chain} is an abstract class with
+two subclasses \ct{Bottom} which denotes an empty sequence and \ct{Knot} which
+denotes a sequence that has a value and a tail sequence that is not evaluated
+yet, providing a \ct{BlockClosure} to do that, one item at a time.
 \begin{minted}[fontsize=\footnotesize]{smalltalk}
@@
         yourself
-
+\end{minted}
+Moreover, a \ct{Chain} can be used as argument of a \ct{\#cons:} message send
+\begin{minted}[fontsize=\footnotesize]{smalltalk}
+Chain >> consedObject: car
+    ^ car cons: [self]
+\end{minted}
+directly and other constructors are provided by \ct{Chain} class messages
+\begin{minted}[fontsize=\footnotesize]{smalltalk}
 Chain class >> with: anObj
@@
 \end{minted}
-having two subclasses \ct{Knot} and \ct{Bottom}. A comprehensive
-test case follows \sd{split it and explain.}
+that allow us to write the following comprehensive test case
 \begin{minted}[fontsize=\footnotesize]{smalltalk}
@@
 
-On one hand, when a \ct{Chain} object (i)~receives the message
+On one hand, when a \ct{Chain} object receives the message
 \ct{next:} that carries a \ct{SmallInteger} object $n$ then it keeps at most
@@
 
-On the other hand, when a \ct{Chain} object (ii)~receives the message
+On the other hand, when a \ct{Chain} object receives the message
 \ct{\#contents} it builds a \ct{LinkedList} object \textit{eagerly},
@@
 \begin{minted}[fontsize=\footnotesize]{smalltalk}
-Knot>>#next
+Knot >> next
     ^ linker value
 \end{minted}
-Observe that the \ct{Chain} hierarchy supports the protocol provided
+that allows us to do one more step in filling the not yet evaluated rest of the
+sequence. Observe that the \ct{Chain} hierarchy supports the protocol provided
 by \ct{ReadStream} objects because of the messages \ct{\#next}, \ct{\#next:}
@@
 
-\subsection{\texttt{Runner}s, \texttt{Reifier}s and \texttt{Var}s}
+\subsection{The Runner}
 
-Recall \sd{it was not said that way so I do not recall} that a \ct{Goal} object responds with a \ct{Chain} that contains either zero,
-one, or more substitutions when it receives the message \ct{\#onState:} and it
-delegates the task to process those substitutions to \ct{Runner} objects,
+We can now be more precise and rephrase that a \ct{Goal} object responds with a
+\ct{Chain} that contains either zero, one or more substitutions when it
+receives the message \ct{\#onState:}.
 
+In this section we describe three components that have the responsibility to
+report and to present the substitutions that satisfy the given goal. The entry
+point is the message
 \begin{minted}[fontsize=\footnotesize]{smalltalk}
@@
 \end{minted}
-which are able to process both (i)~all of them with the message
+that allows a goal to build a \ct{Runner} which has the responsibilities to
+process the \ct{Chain} of substitutions; in particular, (i)~all of them with
+the message
 \begin{minted}[fontsize=\footnotesize]{smalltalk}
@@
 \end{minted}
-and (ii)~at most $n$ of them with the message \ct{atMost:} below: 
+and (ii)~at most $n$ of them with the message
 \begin{minted}[fontsize=\footnotesize]{smalltalk}
@@
 \end{minted}
-relying on the unary message \ct{run}
 
+Those messages rely on the unary message
 \begin{minted}[fontsize=\footnotesize]{smalltalk}
-Runner>>#run
+Runner >> run
     | states vars selector |
@@
 \end{minted}
-having the responsibilities
+that accomplishes the responsibilities
 \begin{itemize}
 
-\item to send the message \Verb|#onState:| to \Verb|Goal| objects to start the
-resolution process from an empty state, this can be considered the entry point of
-the ``computation'' that produces a \Verb|Chain| of substitutions;
+\item to send the message \Verb|#onState:| to a \Verb|Goal| $g$ in order to
+start the resolution process from an empty state -- this can be considered the
+entry point of the ``computation'' that produces the \Verb|Chain| of
+substitutions that satisfy $g$;
 
 \item to take into account the variables with respect to results should be
-provided; as a particular case, if the root goal doesn't introduce any logic
-variable, by means of the auxiliary messages
+provided; as a particular case, if the root \ct{Goal} $g$ doesn't introduce any logic
+variable, recovered by the messages
 \begin{minted}[fontsize=\footnotesize]{smalltalk}
-Goal>>#vars
+Goal >> vars
     ^ {}
 
-Fresh>>#vars
+Fresh >> vars
     ^ vars
 \end{minted}
-then it is equivalent to prove a \textit{tautology}, represented by
+then it is equivalent to prove that $g$ is a \textit{tautology}, represented
+by
 \begin{minted}[fontsize=\footnotesize]{smalltalk}
-Var class>>#tautology
+Var class >> tautology
     ^ #tautology
 \end{minted}
-which require to decide that a relation is always true regardless values taken
-by logic variables;
+which requires to decide that a relation is always \textit{true} regardless
+the values associated to logic variables;
 
-\item to simplify the results presentation when the root \Verb|Goal| object
-introduces \textit{exactly one} logic variable: in this case a
-\Verb|LinkedList| of values for that variable will be produced; otherwise, a
-\Verb|LinkedList| of \Verb|LinkedList| of values will be produces each one for
-each tuple of logic variables introduced by the root \Verb|Goal| object -- this
-point will be clear when we'll explain \Verb|Fresh| goals that are the only
-ones able to introduce logic variables;
+\item to simplify the presentation of results when the root \Verb|Goal| object
+introduces \textit{exactly one} logic variable $v$: in this case a
+\Verb|LinkedList| of values associated to $v$ will be produced; otherwise, a
+\Verb|LinkedList| of \Verb|LinkedList| of values will be produced, a list for
+each tuple of logic variables introduced -- this point will become clear when
+we'll explain \Verb|Fresh| goals that are the only ones able to introduce logic
+variables;
 
 \item to map each substitution (always translate to \Verb|State| object) that
-satisfies the root \Verb|Goal| object do \textit{reify} the required variables,
-this is accomplished by the messages
-\begin{minted}[fontsize=\footnotesize]{smalltalk}
-s reifier reifyVars: vars
-\end{minted}
-within the block argument of the returning \Verb|#collect:| message send.
+satisfies the root \Verb|Goal| object to a \textit{reified} substitution.
 \end{itemize}
+
 The last point shows that \Verb|Chain| objects are \textit{functors} according
@@
 \begin{minted}[fontsize=\footnotesize]{smalltalk}
-Chain>>#collect: aBlockClosure
+Chain >> collect: aBlockClosure
     ^ self
 
-Knot>>#collect: aBlockClosure
+Knot >> collect: aBlockClosure
     | v |
@@
 \end{minted}
-respectively.  We are now ready to describe the reification process which
+respectively.
+
+\subsection{The Reifier}
+
+We are now ready to describe the reification process which
 requires
 \begin{minted}[fontsize=\footnotesize]{smalltalk}
-State>>#reifier
+State >> reifier
     ^ Reifier on: self
 
-Reifier>>#reifyVars: vars
+Reifier >> reifyVars: vars
     | walker reifier |
@@
                     | w s |
-                    w := walker walkStar: v.
-                    s := reifier reify: w.
-                    s walker walkStar: w]
+                    w := walker value: v.
+                    s := reifier value: w.
+                    s walker value: w]
 \end{minted}
@@
 \begin{minted}[fontsize=\footnotesize]{smalltalk}
-Reifier>>#reify: anObj 
+Reifier >> value: anObj
     ^ (state walk: anObj)
-        reifyUsing: self
+        reifyUsingReifier: self
 \end{minted}
-to dispatch the message \Verb|#reifyUsing:| over
+to dispatch the message \Verb|#reifyUsingReifier:| over
 \begin{minted}[fontsize=\footnotesize]{smalltalk}
-Object>>#reifyUsing: aReifier
+Object >> reifyUsingReifier: aReifier
     ^ aReifier reifyObject: self
 
-Var>>#reifyUsing: aReifier 
-    ^ aReifier reifyVar: self
+Var >> reifyUsingReifier: aReifier
+    ^ aReifier forVar: self
 
-ReifiedVar>>#reifyUsing: aReifier 
-    ^ aReifier reifyReifiedVar: self
+ReifiedVar >> reifyUsingReifier: aReifier
+    ^ aReifier forReifiedVar: self
 
-Array>>#reifyUsing: aReifier
-    ^ aReifier reifyArray: self
+Array >> reifyUsingReifier: aReifier
+    ^ aReifier forArray: self
 
-LinkedList>>#reifyUsing: aReifier 
-    ^ aReifier reifyLinkedList: self
+LinkedList >> reifyUsingReifier: aReifier
+    ^ aReifier forLinkedList: self
 
-Cons>>#reifyUsing: aReifier 
-    ^ aReifier reifyCons: self 
+Cons >> reifyUsingReifier: aReifier
+    ^ aReifier forCons: self
 \end{minted}
@@
 \begin{minted}[fontsize=\footnotesize]{smalltalk}
-Reifier>>#reifyObject: anObj 
+Reifier >> forObject: anObj
     ^ state
 
-Reifier>>#reifyVar: aVar 
+Reifier >> forVar: aVar
     ^ state reifyVar: aVar
 
-Reifier>>#reifyReifiedVar: aVar 
+Reifier >> forReifiedVar: aVar
     ^ state
 
-Reifier>>#reifyArray: aCollection 
-    ^ self reifyCollection: aCollection
+Reifier >> forArray: aCollection
+    ^ self _forCollection: aCollection
 
-Reifier>>#reifyArray: aCollection 
-    ^ self reifyCollection: aCollection
+Reifier >> forLinkedList: aCollection
+    ^ self _forCollection: aCollection
 
-Reifier>>#reifyArray: aCollection 
-    ^ self reifyCollection: aCollection
+Reifier >> forCons: aCollection
+    ^ self _forCollection: aCollection
 
-Reifier>>#reifyCollection: aCollection 
-    ^ aCollection inject: state into: [:s :c | s reifier reify: c]
+Reifier >> _forCollection: aCollection
+    ^ aCollection inject: state into: [:s :c | s reifier value: c]
 \end{minted}
@@
 \begin{minted}[fontsize=\footnotesize]{smalltalk}
-State>>#reifyVar: aVar 
+State >> reifyVar: aVar
     substitution
@@
 \begin{minted}[fontsize=\footnotesize]{smalltalk}
-State>>#walk: anObj 
+State >> walk: anObj
     | k |
@@
 \begin{minted}[fontsize=\footnotesize]{smalltalk}
-Cons>>#inject: init into: aBlock 
+Cons >> inject: init into: aBlock
     | accumulated |
@@
 
+\subsection{The Walker}
 The reification process requires \ct{Walker} objects which are built by
@@
 \begin{minted}[fontsize=\footnotesize]{smalltalk}
-State>>#walker
+State >> walker
     ^ Walker on: self
 \end{minted}
-and it responds to the selector \Verb|#walkStar| with
+and it responds to the message
 \begin{minted}[fontsize=\footnotesize]{smalltalk}
-Walker>>#walkStar: aVar
+Walker >> value: aVar
     ^ (state walk: aVar)
@@
 \begin{minted}[fontsize=\footnotesize]{smalltalk}
-Object>>#walkStarUsingWalker: aWalker
-    ^ aWalker walkStarObject: self
+Object >> walkStarUsingWalker: aWalker
+    ^ aWalker forObject: self
 
-Array>>#walkStarUsingWalker: aWalker
-    ^ aWalker walkStarArray: self
+Array >> walkStarUsingWalker: aWalker
+    ^ aWalker forArray: self
 
-LinkedList>>#walkStarUsingWalker: aWalker
-    ^ aWalker walkStarLinkedList: self
+LinkedList >> walkStarUsingWalker: aWalker
+    ^ aWalker forLinkedList: self
 
-Cons>>#walkStarUsingWalker: aWalker
-    ^ aWalker walkStarCons: self
+Cons >> walkStarUsingWalker: aWalker
+    ^ aWalker forCons: self
 \end{minted}
@@
 \begin{minted}[fontsize=\footnotesize]{smalltalk}
-Walker>>#walkStarObject: anObj
+Walker >> forObject: anObj
     ^ state walk: anObj
 
-Walker>>#walkStarArray: aCollection
+Walker >> forArray: aCollection
+    ^ self _forCollection: aCollection
+
+Walker >> forLinkedList: aCollection
+    ^ self _forCollection: aCollection
+
+Walker >> forCons: aCons
+    ^ self _forCollection: aCons
+
+Walker >> _forCollection: aCollection
     ^ aCollection
-        collect: [:each | self walkStar: each]
-
-Walker>>#walkStarLinkedList: aCollection
-    ^ aCollection
-        collect: [:each | self walkStar: each]
-
-Walker>>#walkStarCons: aCons
-    | car cdr |
-    car := self walkStar: aCons car.
-    cdr := self walkStar: aCons cdr.
-    ^ car cons: cdr
+        collect: [:each | self value: each]
 \end{minted}
@@
 \section{Bootstrapping $\mu$Kanren}
+\label{sec:bootstrapping}
 
@@
 \begin{minted}[fontsize=\footnotesize]{smalltalk}
-GoalTest>>#testFail
+GoalTest >> testFail
     self
@@
 \begin{minted}[fontsize=\footnotesize]{smalltalk}
-False>>#asGoal
+False >> asGoal
     ^ Goal fail
@@
 \begin{minted}[fontsize=\footnotesize]{smalltalk}
-Goal>>#fail
+Goal >> fail
     ^ Fail new
@@
 \begin{minted}[fontsize=\footnotesize]{smalltalk}
-Fail>>#onState: aState
+Fail >> onState: aState
     ^ Chain bottom
@@
 \begin{minted}[fontsize=\footnotesize]{smalltalk}
-GoalTest>>#testSucceed
+GoalTest >> testSucceed
     self
@@
 \begin{minted}[fontsize=\footnotesize]{smalltalk}
-True>>#asGoal
+True >> asGoal
     ^ Goal succeed
@@
 \begin{minted}[fontsize=\footnotesize]{smalltalk}
-Goal>>#succeed
+Goal >> succeed
     ^ Succeed new
@@
 \begin{minted}[fontsize=\footnotesize]{smalltalk}
-Succeed>>#onState: aState
+Succeed >> onState: aState
     ^ Chain with: aState

### 2020-12-04 booklet-tex/introduction.tex
Make room for a new booklet

--- 2019-03-13 booklet/introduction.tex
@@
 Starting small, the next example shows three different
-approaches to define the addition of three \verb|SmallInteger| objects toward 
+approaches to define the addition of three \ct{SmallInteger} objects toward 
 the introduction of the relational paradigm.

### 2020-12-04 booklet/source/conf.py
Add a new folder with some restructuredText content.

@@ -1,0 +1,54 @@
+# Configuration file for the Sphinx documentation builder.
+#
+# This file only contains a selection of the most common options. For a full
+# list see the documentation:
+# https://www.sphinx-doc.org/en/master/usage/configuration.html
+
+# -- Path setup --------------------------------------------------------------
+
+# If extensions (or modules to document with autodoc) are in another directory,
+# add these directories to sys.path here. If the directory is relative to the
+# documentation root, use os.path.abspath to make it absolute, like shown here.
+#
+# import os
+# import sys
+# sys.path.insert(0, os.path.abspath('.'))
+
+
+# -- Project information -----------------------------------------------------
+
+project = 'microkanrenst'
+copyright = '2020, Massimo Nocentini'
+author = 'Massimo Nocentini'
+
+
+# -- General configuration ---------------------------------------------------
+
+# Add any Sphinx extension module names here, as strings. They can be
+# extensions coming with Sphinx (named 'sphinx.ext.*') or your custom
+# ones.
+extensions = [
+        'sphinx.ext.mathjax',
+        'sphinx.ext.githubpages',
+]
+
+# Add any paths that contain templates here, relative to this directory.
+templates_path = ['_templates']
+
+# List of patterns, relative to source directory, that match files and
+# directories to ignore when looking for source files.
+# This pattern also affects html_static_path and html_extra_path.
+exclude_patterns = []
+
+
+# -- Options for HTML output -------------------------------------------------
+
+# The theme to use for HTML and HTML Help pages.  See the documentation for
+# a list of builtin themes.
+#
+html_theme = 'alabaster'
+
+# Add any paths that contain custom static files (such as style sheets) here,
+# relative to this directory. They are copied after the builtin static files,
+# so a file named "default.css" will overwrite the builtin "default.css".
+html_static_path = ['_static']

### 2020-12-07 booklet/source/_ext/pharo.py
First attempt to interface Pharo's comments and definitions with Sphinx.

@@ -1,0 +1,96 @@
+
+import docutils
+from docutils.parsers.rst import Directive
+from docutils.statemachine import ViewList
+from sphinx.domains import Domain
+from sphinx.roles import XRefRole
+from sphinx.util.nodes import nested_parse_with_titles
+from docutils.statemachine import StringList
+import json
+
+# for now hardcore absolute file reference.
+with open('/Users/mn/Developer/snapshots/pharoes/90-microkanren/microkanren.json') as fp:
+    pharoExportDict = json.load(fp)
+
+class PharoDirective(Directive):
+
+    required_arguments = 1
+    optional_arguments = 0
+    final_argument_whitespace = False
+    option_spec = {}
+    has_content = False
+
+    def run(self):
+        classDef = pharoExportDict['classes'][self.arguments[0]]
+
+        rst = StringList()
+
+        dummySourceFilename = '{}.rst'.format(self.arguments[0])
+        for i, l in enumerate(classDef['comment']):
+            rst.append(l, dummySourceFilename, i)
+
+        node = docutils.nodes.section()
+        node.document = self.state.document
+
+        # Parse the rst.
+        nested_parse_with_titles(self.state, rst, node)
+
+        return node.children + [docutils.nodes.literal_block('', classDef['definition'], language='smalltalk')]
+
+class PharoDomain(Domain):
+
+    name = 'pharo'
+    label = 'A Smalltalk domain'
+    roles = {
+        'ref': XRefRole()
+    }
+    directives = {
+        'autoclass': PharoDirective,
+    }
+    indices = {
+    }
+    initial_data = {
+        'classes': [],  # object list
+        #'recipe_ingredients': {},  # name -> object
+    }
+
+    def get_full_qualified_name(self, node):
+        return '{}.{}'.format('pharo', node.arguments[0])
+
+    def get_objects(self):
+        for obj in self.data['classes']:
+            yield(obj)
+
+    def resolve_xref(self, env, fromdocname, builder, typ, target, node, contnode):
+        match = [(docname, anchor)
+                 for name, sig, typ, docname, anchor, prio
+                 in self.get_objects() if sig == target]
+
+        if len(match) > 0:
+            todocname = match[0][0]
+            targ = match[0][1]
+
+            return make_refnode(builder, fromdocname, todocname, targ, contnode, targ)
+        else:
+            print('Awww, found nothing')
+            return None
+
+    def add_autoclass(self, signature, ingredients):
[…]

### 2020-12-07 booklet/source/conf.py

@@ -10,10 +10,11 @@
 # add these directories to sys.path here. If the directory is relative to the
 # documentation root, use os.path.abspath to make it absolute, like shown here.
 #
-# import os
-# import sys
+import os
+import sys
 # sys.path.insert(0, os.path.abspath('.'))
-
+ 
+sys.path.append(os.path.abspath("./_ext"))
 
 # -- Project information -----------------------------------------------------
 
@@ -30,7 +31,11 @@ author = 'Massimo Nocentini'
 extensions = [
         'sphinx.ext.mathjax',
         'sphinx.ext.githubpages',
+        'pharo',
 ]
+
+# The following configuation values concerns the Pharo domain.
+pharo_json_export_filename='/Users/mn/Developer/snapshots/pharoes/90-microkanren/microkanren.json'
 
 # Add any paths that contain templates here, relative to this directory.
 templates_path = ['_templates']
